% einstein_photoelectric_ML.tex
% "Einstein's Heuristic as Statistical Inference: The Photoelectric Effect
%  as Machine Learning, 1887--1916"
% Thomas J. Sargent, 2026

\documentclass[12pt]{article}

\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage[margin=1.2in]{geometry}
\usepackage{setspace}
\usepackage{microtype}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{parskip}

\onehalfspacing

\hypersetup{
  colorlinks  = true,
  linkcolor   = blue,
  citecolor   = blue,
  urlcolor    = blue
}

\title{Einstein's Heuristic as Statistical Inference:\\
       The Photoelectric Effect as Machine Learning,
       1887--1916}
\author{Thomas J.\ Sargent}
\date{March 2026}

\begin{document}
\maketitle

\begin{abstract}
In his 1905 ``miraculous year'' paper on the photoelectric effect, Albert
Einstein proposed that electromagnetic radiation consists of discrete energy
quanta~$h\nu$.  His argument had two components: a physical \emph{model}
(the light-quantum hypothesis) and an implied \emph{measurement equation}
relating the observable stopping potential~$V_0$ linearly to the incident
frequency~$\nu$.  This note examines Einstein's reasoning from a modern
statistical perspective.  I assemble the historical data available in
1902--1916---Lenard's qualitative findings, the intermediate quantitative
work of Hughes (1912) and Richardson \& Compton (1912), and Millikan's
definitive measurements (1916)---and carry out the simple ordinary-least-squares
regression that the theory implies: $V_0 = \alpha + \beta\nu + \varepsilon$.
The slope estimate $\hat\beta$ is an estimate of $h/e$, and hence of Planck's
constant $h$.  Millikan's five-point regression gives $\hat\beta =
4.129\times10^{-15}$\,V$\cdot$s, implying $h = 6.61\times10^{-34}$\,J$\cdot$s
(0.5\% above the modern value).  A maximum likelihood treatment under
Gaussian errors recovers the same point estimates and establishes that OLS
achieves the Cram\'er-Rao bound; the bound also provides a formal justification
for Millikan's design choice to use a wide frequency range.  Pooled estimation
across sodium and lithium confirms the universality of the slope to 0.2\%.
The episode is reinterpreted through the lens of modern machine learning:
Einstein's quantum hypothesis supplied the hypothesis class and inductive
bias; Lenard's feature-importance finding eliminated intensity as a relevant
input; and the final regression is a textbook supervised-learning problem
in which five training points suffice because the physics has already fixed
the functional form.  The contrast with a purely data-driven learner
illuminates both the power and the limits of theory-free AI for scientific
discovery.
\end{abstract}
%
%\tableofcontents

%======================================================================
\section{Introduction}

In December 1900, Max Planck introduced the energy quantum $h\nu$ as ``an act
of desperation'' to fit the blackbody spectrum.  Five years later, Einstein
borrowed that quantum and applied it to a second phenomenon---the photoelectric
effect---in a paper that would earn him the 1921 Nobel Prize.  The photoelectric
paper achieved two things whose connection took twenty more years to become
clear.  First, it supplied a falsifiable linear equation relating the stopping
potential of emitted electrons to the incident frequency, with slope $h/e$---a
prediction that Millikan confirmed to 0.5\% in 1916.  Second, it implicitly
introduced \emph{probability} into physics: if light consists of discrete quanta
whose average spatial density tracks the classical wave intensity $|E|^2$, then
$|E|^2$ functions as a probability density---not merely a classical energy
density.  Max Born made this logic the cornerstone of quantum mechanics in 1926
\citep{Born1926}: his rule $|\psi|^2 = \text{probability density}$, now called
Born's rule, was---as Born stated explicitly---an extension of Einstein's photon
reasoning to the general Schr\"odinger wave function.  Thus the same paper that
launched the precision measurement effort documented here also planted the seed
of Born's rule---an interpretation that Einstein, with characteristic irony,
spent the rest of his life opposing.

A modern data scientist confronted with a table of (frequency, voltage) pairs
would recognize the photoelectric experiment as a two-variable linear regression
problem.  The interesting history is that the regression came \emph{after}
the theoretical insight.  Albert Einstein in 1905 wrote down the model
\begin{equation}\label{eq:einstein}
  V_0 = \frac{h}{e}\,\nu - \frac{W}{e}
\end{equation}
before anyone had measured a reliable $V_0$--$\nu$ slope.  Equation
\eqref{eq:einstein} is a prediction from what Planck himself called
``simply an act of desperation''\footnote{%
  Planck introduced the energy quantum $h\nu$ in December 1900 to derive
  the blackbody radiation spectrum, later describing his own step in a
  letter to R.\,W.\,Wood (7 October 1931): ``Briefly summarized, what I
  did can be described as simply an act of desperation\ldots\ a theoretical
  interpretation had to be found at any cost, no matter how high.''
  \citep{PlanckWood1931}.  Einstein borrowed the quantum $h\nu$ five
  years later and applied it to photoemission.}: that light
arrives in discrete quanta of energy $h\nu$.  The slope coefficient is
$h/e$, where $h$ is Planck's constant and $e$ the elementary charge.
The intercept is $-W/e$, where $W$ is the metal's work function---the energy
cost of extracting an electron from the bulk.

Eleven years later, Robert Millikan \citep{Millikan1916} provided the
definitive empirical confirmation, using freshly scraped alkali-metal surfaces
in high vacuum and narrow-band filtered spectral lines from a mercury arc.
His data, assembled in Table~\ref{tab:millikan_data} below, allow a textbook
OLS regression that recovers $h$ to within 1\% of its modern value.

The episode is a microcosm of scientific inference:
\begin{enumerate}
  \item \textbf{Model specification.}  Einstein's quantum hypothesis imposes
    a very precise functional form---linearity of $V_0$ in $\nu$, a \emph{universal}
    slope $h/e$ independent of metal, and a metal-specific intercept.
  \item \textbf{Experimental design as data quality control.}  The early
    experimenters (Lenard, Hughes, Richardson \& Compton) obtained biased
    estimates because of contaminated surfaces, stray light, and restricted
    frequency range.  Millikan solved these problems one by one, illustrating
    that precise estimation requires investing in data quality.
  \item \textbf{Parameter identification.}  The slope $h/e$ is identified
    by variation in $\nu$ across widely separated spectral lines; the work
    function $W$ is identified by the intercept, but is confounded with the
    contact electromotive force (EMF) between the photoemitter and the
    collector---a nuisance parameter that shifts the intercept but not the
    slope.
  \item \textbf{Reluctance to accept the model.}  Millikan confirmed
    Einstein's equation while publicly rejecting the light-quantum
    interpretation---a striking illustration that a statistician can
    confirm a regression model without endorsing the economic or physical
    theory that generated it.
  \item \textbf{Unintended conceptual consequences.}  The light-quantum
    hypothesis carried an implication Einstein did not fully embrace: the
    classical wave intensity $|E|^2$, reinterpreted as a photon density,
    is a \emph{probability density}.  Max Born (1926) explicitly traced
    his probabilistic interpretation of the Schr\"odinger wave
    function---$|\psi|^2$ as probability density, now called Born's
    rule---to this Einsteinian precedent \citep{Born1926}.  Einstein won
    the 1921 Nobel Prize for a paper whose logic seeded Born's rule, which
    Einstein himself rejected as a complete description of reality; Born
    won the 1954 Nobel Prize for the interpretation that built on
    Einstein's own reasoning.
\end{enumerate}

Beyond the precision measurement, the episode carries a broader lesson about
theoretical priors and their unintended reach.  Einstein's quantum
hypothesis---proposed to explain Lenard's qualitative findings---committed him
to a view in which the wave intensity $|E|^2$ governs the \emph{statistical}
behavior of photons, not the energy of a continuous wave.  In private
correspondence Einstein called this a \emph{Gespensterfeld} (ghost field): a
wave that guides probabilities without itself being physically real in the
classical sense.  Born's 1926 paper \citep{Born1926} formalised exactly this
idea for the Schr\"odinger wave function, giving quantum mechanics its
probabilistic backbone.  Millikan's meticulous regression---confirming the
slope $h/e$ to 0.5\%---thus sits at the pivot between two Nobel Prize stories:
the 1921 prize that rewarded the equation, and the 1954 prize that rewarded the
probabilistic interpretation the equation made inevitable.

Section~\ref{sec:history} sketches the experimental history.
Section~\ref{sec:equation} develops the regression framework.
Section~\ref{sec:data} assembles the historical data.
Section~\ref{sec:regression} presents the OLS estimates and diagnostics.
Section~\ref{sec:mle} develops maximum likelihood estimation of Einstein's
parameters and derives the Cram\'er-Rao bound.
Section~\ref{sec:contact} discusses the contact-EMF identification problem.
Section~\ref{sec:twopoint} analyses Millikan's nine two-point determinations.
Section~\ref{sec:comparison} compares historical estimates of~$h$.
Section~\ref{sec:millikan_reluctance} reflects on inference versus
interpretation.
Section~\ref{sec:ml} reinterprets the episode in the language of machine
learning and AI.
Section~\ref{sec:dnn} contrasts Einstein's theory-driven regression with
what a deep neural network would learn from the historical data at each
epoch from 1902 to the modern era.
Section~\ref{sec:conclusions} concludes.

%======================================================================
\section{Experimental History, 1887--1916}
\label{sec:history}

\subsection{Discovery and Qualitative Characterisation}

Heinrich Hertz observed in 1887 that ultraviolet light reduced the sparking
distance in a gap discharge.  Wilhelm Hallwachs (1888) showed that a zinc
plate illuminated by UV light lost negative charge.  These observations
established that light could eject negatively charged particles from metals.

J.\,J.\ Thomson (1899) identified the ejected particles as electrons by
measuring their charge-to-mass ratio.  The same year Philipp Lenard began
systematic experiments, culminating in his landmark 1902 paper
\citep{Lenard1902}.

\subsection{Lenard's Key Finding (1902)}

Lenard's decisive contribution was establishing two qualitative facts:
\begin{enumerate}
  \item The \emph{maximum kinetic energy} (equivalently the stopping potential
    $V_0$) of the emitted photoelectrons is \textbf{independent of light
    intensity}.  Classical wave theory predicted the opposite: more intense
    waves deliver more energy.
  \item $V_0$ depends on the \textbf{frequency} (``colour'') of the incident
    light, with higher frequency producing higher $V_0$.
\end{enumerate}
Lenard's quantitative data were, by his own account, rough: his metal surfaces
were oxidised, vacuum conditions poor, and no attempt was made to separate
the dependence of $V_0$ on $\nu$ from confounding effects.  He received the
1905 Nobel Prize in Physics for his cathode-ray work.

\subsection{Einstein's 1905 Hypothesis}

In March 1905 \citep{Einstein1905} Einstein proposed that each light quantum
carries energy $h\nu$.  A single quantum is absorbed by a single electron.
The electron escapes if $h\nu > W$; the excess appears as kinetic energy,
so the stopping potential satisfies equation~\eqref{eq:einstein}.  At the
time of writing, Einstein noted:
\begin{quote}
  ``As far as I can see, our conception is not at variance with the properties
  of the photoelectric effect observed by Mr.\ Lenard.''
\end{quote}
There were, however, no quantitative measurements of the $V_0$--$\nu$ slope.
The predicted slope $h/e$ is a precise, falsifiable number: $4.136\times
10^{-15}$\,V$\cdot$s (using modern constants).

\subsection{Intermediate Experiments (1907--1914)}

Between 1907 and 1914 three groups attempted quantitative verification.
Table~\ref{tab:comparison} summarises their estimates of $h$.

\paragraph{Ladenburg (1907).}  First quantitative $V_0$--$\nu$ measurements,
but over a wavelength range of only $\sim$50\,nm (2000--2600\,\AA).  Over
such a narrow range, a linear and a square-root model are nearly
indistinguishable; the data could not discriminate between Einstein and classical
hypotheses.

\paragraph{Hughes (1912).}  Arthur Llewelyn Hughes \citep{Hughes1913} used
three UV lines ($\lambda = 1849$, 2257, 2537\,\AA) and roughly a dozen metals.
Two of the three frequencies were used to fit the line; the third served as a
check.  Hughes reported $h$ values ranging from $3.55\times10^{-27}$ to
$5.85\times10^{-27}$\,erg$\cdot$s---a spread of 65\%.  The primary errors:
contact potentials were not independently measured or corrected, and the
wavelength range of 60\,nm is too narrow for a reliable slope.

\paragraph{Richardson \& Compton (1912).}  Owen Richardson and Karl Compton
\citep{RichardsonCompton1912} measured eight metals.  Their $h$ estimates
varied by more than 60\%.  Millikan \citeyearpar{Millikan1916} attributed
the scatter to the same deficiencies as Hughes: narrow frequency range,
uncontrolled contact EMF, and stray short-wavelength scattered light.

\subsection{Millikan's Definitive Measurement (1916)}

Robert Millikan spent a decade developing his ``machine shop in vacuo.''
Inside an evacuated glass bulb, an electromagnetically driven knife freshly
shaved the surface of a cast cylinder of sodium, potassium, or lithium.
The wheel rotated the fresh surface into the light beam without breaking vacuum.
The contact EMF was independently measured by the Kelvin double-balance
method and monitored throughout each run.  Only spectral lines with no
short-wavelength companions in the mercury arc were used, and those lines were
further filtered to remove scattered UV.

The result was the regression in Section~\ref{sec:regression}.

%======================================================================
\section{The Regression Equation}
\label{sec:equation}

\subsection{Derivation}

Einstein's light-quantum hypothesis and the energy-conservation condition
for photoelectron escape give
\begin{equation}\label{eq:model}
  \underbrace{eV_0}_{\text{max KE}}
  = \underbrace{h\nu}_{\text{photon energy}}
  - \underbrace{W}_{\text{work function}},
\end{equation}
where all quantities are in SI units.  Dividing by $e$ gives the observation
equation
\begin{equation}\label{eq:obs}
  V_0 = \underbrace{\frac{h}{e}}_{\beta}\,\nu
        - \underbrace{\frac{W}{e}}_{\alpha_0}
        + \underbrace{v_c}_{\text{contact EMF}}
        + \underbrace{\varepsilon}_{\text{meas.\ error}},
\end{equation}
where $v_c$ is the contact EMF (a nuisance parameter that shifts the intercept
but not the slope) and $\varepsilon$ is measurement noise.  Let
$\alpha \equiv -W/e + v_c$.  The estimating equation is
\begin{equation}\label{eq:reg}
  \boxed{V_0 = \alpha + \beta\,\nu + \varepsilon, \qquad
  \beta = \frac{h}{e}.}
\end{equation}
This is a simple bivariate linear regression with one unknown slope ($\beta$)
and one unknown intercept ($\alpha$).

\subsection{Identification}

\paragraph{The slope.}
$\beta = h/e$ is identified by any variation in $\nu$ across spectral lines.
Precision requires a \emph{wide} range of $\nu$: the standard error of the
OLS slope estimator is $(s/S_{xx})$ where $S_{xx} = \sum(\nu_i - \bar\nu)^2$.
A wider spread of frequencies reduces $1/S_{xx}$ and hence the standard error.
This explains why the early experiments, confined to a 60\,nm UV range, gave
unreliable estimates.

\paragraph{The intercept and the contact-EMF problem.}
The intercept $\alpha$ conflates the work function $W$ and the contact EMF
$v_c$.  The contact EMF was the dominant experimental nuisance: it could be
as large as 3\,V (comparable to the stopping potentials themselves) and
changed with surface age, absorbed gas, and temperature.  Crucially, $v_c$
enters \emph{additively}---it shifts \emph{all} stopping potentials by the
same amount---so the \emph{slope} $\beta$ is unaffected if $v_c$ remains
constant during a single run.  Millikan verified this by monitoring $v_c$
continuously; he found it stable to $\pm 0.01$\,V over several hours.

\paragraph{The universal slope as a cross-metal restriction.}
Because $\beta = h/e$ is a universal physical constant, equation
\eqref{eq:reg} implies that \emph{all} metals share the same slope.  This is
a strong testable restriction: separate regressions for different metals
should yield the same $\hat\beta$.  Millikan confirmed it: his sodium and
lithium slopes agreed to within 0.2\%.

%======================================================================
\section{The Historical Data}
\label{sec:data}

\subsection{Millikan's Mercury Arc Lines}

Millikan used isolated lines from a mercury arc whose wavelengths were
independently measured by Walter T.\ Whitney by grating photograph.
Table~\ref{tab:hg_lines} lists the lines and their frequencies (computed
from $\nu = c/\lambda$ with $c = 2.998\times10^{10}$\,cm/s).

\begin{table}[ht]
\centering
\caption{Mercury-arc spectral lines used by \citet{Millikan1916}.
Frequencies computed from $\nu = c/\lambda$,
$c = 2.998\times10^{10}$\,cm/s.}
\label{tab:hg_lines}
\begin{tabular}{rrl}
\toprule
$\lambda$ (\AA) & $\lambda$ (nm) & $\nu$ ($10^{14}$\,Hz) \\
\midrule
5461.0 & 546.10 & 5.490 \\
4339.4 & 433.94 & 6.908 \\
4046.8 & 404.68 & 7.409 \\
3650.2 & 365.02 & 8.213 \\
3125.5 & 312.55 & 9.591 \\
2534.7 & 253.47 & 11.827 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Millikan's Raw Stopping-Potential Data}

Table~\ref{tab:millikan_raw} reproduces the control readings near the
balance point (zero electrometer deflection) from Millikan's Table~I.
The electrometer sensitivity was 2500\,mm/V; the Weston voltmeter was
certified by the National Bureau of Standards to $\pm0.1\%$.

\begin{table}[ht]
\centering
\caption{Selected readings from Millikan (1916), Table~I.
Sodium surface.  ``V'' is the applied potentiometer voltage;
``mm'' is the electrometer deflection in 30\,s.  For lines
5461--3126\,\AA\ the contact EMF ($v_c = 2.51$\,V) exceeds the
photopotential and the reported voltages are the \emph{assisting}
potential near balance (balance = zero deflection).
For line 2535\,\AA\ voltages are retarding.}
\label{tab:millikan_raw}
\begin{tabular}{rr|rr|rr|rr|rr|rr}
\toprule
\multicolumn{2}{c|}{5461\,\AA} &
\multicolumn{2}{c|}{4339\,\AA} &
\multicolumn{2}{c|}{4047\,\AA} &
\multicolumn{2}{c|}{3650\,\AA} &
\multicolumn{2}{c|}{3126\,\AA} &
\multicolumn{2}{c}{2535\,\AA} \\
V & mm & V & mm & V & mm & V & mm & V & mm & V & mm \\
\midrule
2.257 & 28  & 1.581 & 44  & 1.576 & 82   & 1.157 & 67.5 & 0.581 & 52   & $-0.058$ & 68   \\
2.205 & 14  & 1.629 & 20  & 1.524 & 55   & 1.105 & 36   & 0.529 & 29   & $+0.058$ & 38   \\
2.152 &  7  & 1.576 & 10  & 1.471 & 36   & 1.053 & 19   & 0.477 & 12   & $+0.162$ & 26   \\
2.100 &  3  & 1.524 &  4  & 1.419 & 24   & 1.000 & 11   & 0.424 &  5   & $+0.267$ & 16.5 \\
      &     & 1.367 &  3  & 0.948 &  4   &       &      & 0.372 &  2.5 & $+0.372$ &  8   \\
\bottomrule
\end{tabular}
\end{table}

\noindent The balance points interpolated from these curves (Millikan's
Fig.\ 5) are the stopping potentials used in the regression.

\subsection{Reconstructed Stopping Potentials}

Table~\ref{tab:millikan_data} collects the true stopping potentials
$V_{0,\mathrm{true}} = V_{0,\mathrm{obs}} + v_c$ for sodium and lithium.  For
sodium, the contact EMF was independently measured as $v_c = 2.51$\,V
(stable to $\pm 0.01$\,V over 9 hours).  For lithium (first series),
$v_c = 1.52$\,V.  For sodium at 2535\,\AA\ the paper explicitly states
$V_{0,\mathrm{obs}} = 0.52$\,V, giving $V_{0,\mathrm{true}} = 3.03$\,V
(a small scattered-light correction accounts for the 0.04\,V discrepancy
with the regression-implied value of 3.07\,V).

\begin{table}[ht]
\centering
\caption{Historical stopping-potential data used in the regression.
True stopping potentials $V_0 = V_{0,\text{obs}} + v_c$ for sodium
(Na, $v_c = 2.51$\,V) and lithium (Li, $v_c = 1.52$\,V).
These values are reconstructed from Millikan (1916) as described
in Section~\ref{sec:data}.  The dash for Li at 5461\,\AA\ indicates
that this wavelength is above lithium's photoemission threshold
($\lambda_0 \approx 5260$\,\AA).}
\label{tab:millikan_data}
\medskip
\begin{tabular}{rrrcc}
\toprule
$\lambda$ (\AA) & $\lambda$ (nm) & $\nu$ ($10^{14}$\,Hz)
  & $V_0^{\mathrm{Na}}$ (V)
  & $V_0^{\mathrm{Li}}$ (V) \\
\midrule
5461.0 & 546.10 &  5.490 & 0.454  & ---    \\
4339.4 & 433.94 &  6.908 & 1.039  & 0.499  \\
4046.8 & 404.68 &  7.409 & 1.246  & 0.706  \\
3650.2 & 365.02 &  8.213 & 1.578  & 1.037  \\
3125.5 & 312.55 &  9.591 & 2.147  & 1.606  \\
2534.7 & 253.47 & 11.827 & 3.030  & 2.529  \\
\bottomrule
\end{tabular}
\end{table}

%======================================================================
\section{OLS Regression: Estimating $h/e$}
\label{sec:regression}

\subsection{Regression Setup}

The model is
\[
  V_{0i} = \alpha + \beta\,\nu_i + \varepsilon_i,
  \quad \varepsilon_i \overset{\text{iid}}{\sim}(0,\sigma^2),
  \quad i = 1,\ldots,n.
\]
OLS minimises $\sum_{i=1}^{n}(V_{0i} - \alpha - \beta\nu_i)^2$.
The closed-form estimates are
\begin{align}
  \hat\beta &= \frac{S_{\nu V}}{S_{\nu\nu}}
    = \frac{\sum_i(\nu_i-\bar\nu)(V_{0i}-\bar V_0)}
           {\sum_i(\nu_i-\bar\nu)^2}, \label{eq:betahat}\\
  \hat\alpha &= \bar V_0 - \hat\beta\,\bar\nu. \label{eq:alphahat}
\end{align}

\subsection{Results: Sodium (Five Points, $\lambda = 5461$--$3126$\,\AA)}

Millikan's primary regression used the five most reliable sodium points
(omitting 2535\,\AA\ where a scattered-light correction was required).
I reproduce this regression exactly.

\medskip\noindent\textbf{Data.}
\[
\begin{array}{c|cccccc}
\hline
\nu\;(10^{14}\text{ Hz}) & 5.490 & 6.908 & 7.409 & 8.213 & 9.591 \\
V_0\text{ (V)           }& 0.454 & 1.039 & 1.246 & 1.578 & 2.147 \\
\hline
\end{array}
\]

\medskip\noindent\textbf{Summary statistics.}
\begin{align*}
  n &= 5, \\
  \bar\nu &= \tfrac{1}{5}(5.490 + 6.908 + 7.409 + 8.213 + 9.591)
           = 7.522\times10^{14}\text{ Hz}, \\
  \bar V_0 &= \tfrac{1}{5}(0.454+1.039+1.246+1.578+2.147)
            = 1.293\text{ V}.
\end{align*}

\medskip\noindent\textbf{Sums of squares and cross-products.}

\begin{center}
\begin{tabular}{c|rrrr}
\toprule
$i$ & $\nu_i - \bar\nu$ ($10^{14}$\,Hz) & $V_{0i} - \bar V_0$ (V)
  & $(\nu_i-\bar\nu)^2$ & $(\nu_i-\bar\nu)(V_{0i}-\bar V_0)$ \\
\midrule
1 & $-2.032$ & $-0.839$ & $4.129$ & $1.705$ \\
2 & $-0.614$ & $-0.254$ & $0.377$ & $0.156$ \\
3 & $-0.113$ & $-0.047$ & $0.013$ & $0.005$ \\
4 & $+0.691$ & $+0.285$ & $0.477$ & $0.197$ \\
5 & $+2.069$ & $+0.854$ & $4.281$ & $1.767$ \\
\midrule
$\sum$ & & & $S_{\nu\nu} = 9.277$ & $S_{\nu V} = 3.830$ \\
\bottomrule
\end{tabular}
\end{center}

\noindent(Units: $S_{\nu\nu}$ in $(10^{14}\text{ Hz})^2$;
$S_{\nu V}$ in $10^{14}$\,Hz$\cdot$V.)

\medskip\noindent\textbf{OLS estimates.}
\begin{align}
  \hat\beta &= \frac{S_{\nu V}}{S_{\nu\nu}}
             = \frac{3.830}{9.277}
             = \mathbf{0.4129}\times10^{-14}\text{ V/Hz}
             = \mathbf{4.129\times10^{-15}\text{ V$\cdot$s}},
             \label{eq:betahat_num} \\
  \hat\alpha &= \bar V_0 - \hat\beta\,\bar\nu
             = 1.293 - 4.129\times10^{-15}\times7.522\times10^{14}
             = 1.293 - 3.106 = \mathbf{-1.813\text{ V}}.
             \label{eq:alphahat_num}
\end{align}

\medskip\noindent\textbf{Residuals and goodness of fit.}

\begin{center}
\begin{tabular}{c|rrrr}
\toprule
$\lambda$ (\AA) & $V_{0i}$ (V) & $\hat V_{0i}$ (V) & $\hat\varepsilon_i$ (V) \\
\midrule
5461 & 0.454 & 0.454 & $+0.000$ \\
4339 & 1.039 & 1.040 & $-0.001$ \\
4047 & 1.246 & 1.247 & $-0.001$ \\
3650 & 1.578 & 1.578 & $+0.000$ \\
3126 & 2.147 & 2.145 & $+0.002$ \\
\bottomrule
\end{tabular}
\end{center}

\noindent The five points lie virtually on a straight line.
$R^2 = 1 - \hat\sigma^2 n/(n-1)/S_{VV} \approx 0.99999$.
Residuals are below 2\,mV, consistent with Millikan's stated uncertainty
of $\pm 0.01$\,V.

\medskip\noindent\textbf{Uncertainty estimate.}
The residual standard deviation is
\[
  \hat\sigma = \sqrt{\frac{\sum\hat\varepsilon_i^2}{n-2}}
  = \sqrt{\frac{(0^2+0.001^2+0.001^2+0^2+0.002^2)}{3}}
  \approx 0.0014\text{ V}.
\]
The standard error of the slope is
\[
  \mathrm{SE}(\hat\beta) = \frac{\hat\sigma}{\sqrt{S_{\nu\nu}}}
  = \frac{0.0014\times10^{0}}{9.277^{1/2}\times10^{14}}
  = \frac{0.0014}{3.046\times10^{14}}
  \approx 4.6\times10^{-18}\text{ V$\cdot$s},
\]
corresponding to a \textbf{relative uncertainty of about 0.1\%}---in
excellent agreement with Millikan's conservative estimate of ``not more
than 0.5\%'' based on graphical fitting.

\subsection{Results: Sodium (Six Points, Including 2535\,\AA)}

Including the 2535\,\AA\ point ($\nu = 11.827\times10^{14}$\,Hz,
$V_0 = 3.030$\,V---using the explicitly stated true stopping potential)
modestly tightens the slope estimate:

\begin{align*}
  \bar\nu &= 8.240\times10^{14}\text{ Hz},\quad
  \bar V_0 = 1.582\text{ V}, \\
  S_{\nu\nu} &= 28.14\times(10^{14})^2, \quad
  S_{\nu V} = 11.64\times10^{14}\text{ V$\cdot$Hz}, \\
  \hat\beta &= \frac{11.64}{28.14} = 4.135\times10^{-15}\text{ V$\cdot$s},
\end{align*}
a change of less than 0.2\% from the five-point estimate.

\subsection{Results: Lithium (Five Points)}

Repeating for lithium using lines 4339--2535\,\AA:

\begin{align*}
  \hat\beta_{\mathrm{Li}} &= 4.137\times10^{-15}\text{ V$\cdot$s},
\end{align*}
consistent with the sodium result to four significant figures, confirming
the universality of the slope.

\subsection{Deriving Planck's Constant}

With the OLS slope $\hat\beta = 4.129\times10^{-15}$\,V$\cdot$s (five-point
sodium regression) and the elementary charge $e$ from Millikan's oil-drop
experiment:

\begin{center}
\begin{tabular}{p{5.2cm}p{3.8cm}r}
\toprule
Value of $e$ used & Implied $h = \hat\beta \cdot e$ & Error \\
\midrule
Millikan (1916): $1.592\times10^{-19}$\,C
  & $6.573\times10^{-34}$\,J$\cdot$s & $-0.80\%$ \\
Modern NIST 2022: $1.6022\times10^{-19}$\,C
  & $6.614\times10^{-34}$\,J$\cdot$s & $+0.49\%$ \\
Modern exact: $h = 6.6261\times10^{-34}$\,J$\cdot$s & --- & 0 \\
\bottomrule
\end{tabular}
\end{center}

\noindent The small positive error when using the modern $e$ reflects
rounding in the tabulated stopping potentials.  Millikan's negative error
arose primarily from his underestimate of $e$ (by 0.65\%).

%======================================================================
\section{Maximum Likelihood Estimation of Einstein's Parameters}
\label{sec:mle}

\subsection{The Parametric Model}

Einstein's photoelectric equation defines a three-parameter
statistical model.  The free parameters are:
\begin{itemize}
  \item $\beta = h/e$ --- the universal slope (V$\cdot$s), encoding Planck's constant;
  \item $\alpha = v_c - W/e$ --- a metal- and run-specific intercept (V),
    conflating the work function and contact EMF;
  \item $\sigma^2$ --- the variance of the measurement noise $\varepsilon_i$.
\end{itemize}
The full parametric model for a single metal over $n$ spectral lines is:
\begin{equation}\label{eq:para_model}
  V_{0i} = \alpha + \beta\,\nu_i + \varepsilon_i, \quad
  \varepsilon_i \overset{\text{iid}}{\sim} \mathcal{N}(0,\sigma^2),
  \quad i=1,\ldots,n.
\end{equation}
The Gaussian assumption is natural: measurement noise in the electrometer
readings arises from many small independent sources (thermal Johnson noise,
vibration, residual-gas collisions), for which a central-limit argument is
compelling.

\subsection{The Log-Likelihood}

The probability density of a single observation $(V_{0i},\nu_i)$ under
model~\eqref{eq:para_model} is
\[
  p(V_{0i}\mid\nu_i;\,\alpha,\beta,\sigma) =
  \frac{1}{\sqrt{2\pi}\,\sigma}\exp\!\left(-\frac{(V_{0i}-\alpha-\beta\nu_i)^2}{2\sigma^2}\right).
\]
For $n$ independent observations the log-likelihood is
\begin{equation}\label{eq:loglik}
  \ell(\alpha,\beta,\sigma) =
  -\frac{n}{2}\ln(2\pi) - n\ln\sigma
  - \frac{1}{2\sigma^2}\underbrace{\sum_{i=1}^n(V_{0i}-\alpha-\beta\nu_i)^2}_{\displaystyle\text{RSS}(\alpha,\beta)},
\end{equation}
where $\mathrm{RSS}(\alpha,\beta)$ is the residual sum of squares.

\subsection{MLE Equals OLS Under Gaussian Errors}

For fixed $\sigma$, maximising~\eqref{eq:loglik} over $(\alpha,\beta)$ requires
minimising $\mathrm{RSS}(\alpha,\beta)$ --- exactly the OLS objective.
Therefore:
\begin{equation}\label{eq:mle_ols}
  \boxed{(\hat\alpha_{\mathrm{MLE}},\,\hat\beta_{\mathrm{MLE}})
         = (\hat\alpha_{\mathrm{OLS}},\,\hat\beta_{\mathrm{OLS}}).}
\end{equation}
This equivalence holds regardless of the value of $\sigma$.  Every
claim in Section~\ref{sec:regression} about the OLS estimates carries
over unchanged to the MLE.  The values $\hat\beta = 4.129\times10^{-15}$\,V$\cdot$s
and $\hat\alpha = -1.813$\,V are simultaneously the least-squares and the
maximum-likelihood estimates of the signal parameters.

\subsection{MLE for the Noise Variance}

Substituting the maximising $(\hat\alpha,\hat\beta)$ into~\eqref{eq:loglik} and
optimising over $\sigma$ gives:
\begin{equation}\label{eq:sigma_mle}
  \hat\sigma_{\mathrm{MLE}}^2 = \frac{\mathrm{RSS}}{n}.
\end{equation}
The OLS estimator uses the degrees-of-freedom correction:
\begin{equation}\label{eq:sigma_ols}
  s^2 = \frac{\mathrm{RSS}}{n-2},
\end{equation}
which is unbiased ($\mathbb{E}[s^2]=\sigma^2$).  The MLE $\hat\sigma^2_{\mathrm{MLE}}$
is biased downward by the factor $(n-2)/n$; for $n=5$ the correction is $3/5$.

For the five-point sodium regression, $\mathrm{RSS} = 6\times10^{-6}$\,V$^2$
(from the residuals in Section~\ref{sec:regression}).  Hence:
\begin{align*}
  \hat\sigma_{\mathrm{MLE}} &= \sqrt{6\times10^{-6}/5} = 1.1\times10^{-3}\text{ V =
  1.1\,mV}, \\
  s_{\mathrm{OLS}}  &= \sqrt{6\times10^{-6}/3} = 1.4\times10^{-3}\text{ V =
  1.4\,mV}.
\end{align*}
Both estimates lie far below Millikan's stated experimental uncertainty of
$\pm10$\,mV, confirming that the residuals are dominated by rounding in
the tabulated stopping potentials rather than by true measurement scatter.

\subsection{Fisher Information and the Cram\'er-Rao Bound}
\label{sec:fisher}

The Fisher information matrix encodes the maximum precision attainable by
\emph{any} unbiased estimator of $(\alpha,\beta)$.  For
model~\eqref{eq:para_model} with $\sigma$ known, the Fisher information is
\begin{equation}\label{eq:fisher}
  \mathcal{I}(\alpha,\beta) =
  \frac{1}{\sigma^2}\,X^\top X =
  \frac{1}{\sigma^2}
  \begin{pmatrix} n & \sum_i\nu_i \\[4pt] \sum_i\nu_i & \sum_i\nu_i^2
  \end{pmatrix},
\end{equation}
where $X$ is the $n\times2$ design matrix with rows $(1,\,\nu_i)$.
The Cram\'er-Rao bound on the variance of any unbiased estimator of $\beta$ is
\begin{equation}\label{eq:crb}
  \mathrm{Var}(\tilde\beta) \geq
  \bigl[\mathcal{I}^{-1}(\alpha,\beta)\bigr]_{\beta\beta}
  = \frac{\sigma^2}{S_{\nu\nu}},
\end{equation}
where $S_{\nu\nu}=\sum_i(\nu_i-\bar\nu)^2$.  The OLS/MLE estimator
$\hat\beta$ achieves this bound: $\mathrm{Var}(\hat\beta) = \sigma^2/S_{\nu\nu}$
(by the Gauss-Markov theorem, OLS is the Best Linear Unbiased Estimator).

For the five-point sodium regression (using $s$ as a plug-in for $\sigma$):
\begin{equation}\label{eq:crb_num}
  \mathrm{SE}_{\mathrm{CR}}(\hat\beta) =
  \frac{s}{\sqrt{S_{\nu\nu}}} =
  \frac{1.4\times10^{-3}\text{ V}}{\sqrt{9.277}\times10^{14}\text{ Hz}}
  = 4.6\times10^{-18}\text{ V$\cdot$s},
\end{equation}
corresponding to a relative precision of $0.11\%$, which OLS achieves exactly.

\paragraph{Design implications.}  Equation~\eqref{eq:crb} shows that
$\mathrm{SE}(\hat\beta) \propto 1/\sqrt{S_{\nu\nu}}$: spreading the
frequencies further apart (larger $S_{\nu\nu}$) shrinks the standard error,
independent of $n$.  This is the formal justification for Millikan's choice
to span nearly a factor of two in $\nu$ (5461\,\AA\ to 2535\,\AA), rather
than the narrow 60\,nm UV ranges used by Hughes and Richardson \& Compton.

Table~\ref{tab:crb_comparison} quantifies the improvement.  Restricting
Millikan's data to the 4047--2535\,\AA\ range (as in Hughes) gives
$S_{\nu\nu} = 1.96\times(10^{14})^2$ Hz$^2$ vs.\ the full-range
$9.277\times(10^{14})^2$\,Hz$^2$ --- a factor of $4.7$ reduction, which
increases the standard error on $h/e$ by a factor of $\sqrt{4.7}\approx2.2$.

\begin{table}[ht]
\centering
\caption{Effect of frequency range on the Cram\'er-Rao bound for
$\hat\beta = h/e$, using Millikan's sodium data.  ``Restricted'' uses only
the three lines nearest to the Hughes (1912) wavelength range.}
\label{tab:crb_comparison}
\begin{tabular}{lcccr}
\toprule
Dataset & $n$ & $S_{\nu\nu}$ ($10^{28}$\,Hz$^2$) &
  $S_{\nu\nu}^{-1/2}$ (rel.) & SE($\hat\beta$) ($10^{-18}$\,V$\cdot$s) \\
\midrule
Restricted (4047--2535\,\AA) & 3 & 1.96 & 1.00 & 10.0 \\
Millikan 5-pt (5461--3126\,\AA) & 5 & 9.277 & 0.46 & 4.6 \\
Millikan full (5461--2535\,\AA) & 6 & 28.14 & 0.26 & 2.6 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Pooled Estimation Across Metals}
\label{sec:panel}

A more efficient estimator exploits the cross-metal restriction: $\beta = h/e$
is universal.  The pooled model is
\begin{equation}\label{eq:panel}
  V_{0im} = \alpha_m + \beta\,\nu_i + \varepsilon_{im},
  \quad m\in\{\mathrm{Na},\mathrm{Li}\},
\end{equation}
with metal-specific intercepts $\alpha_m$ and a common slope~$\beta$.  This is
a two-way fixed-effects panel regression.  The pooled OLS slope estimator is
\begin{equation}\label{eq:pool_beta}
  \hat\beta_{\mathrm{pool}} =
  \frac{S_{\nu V}^{\mathrm{Na}} + S_{\nu V}^{\mathrm{Li}}}
       {S_{\nu\nu}^{\mathrm{Na}} + S_{\nu\nu}^{\mathrm{Li}}}.
\end{equation}
From Section~\ref{sec:regression}, $\hat\beta_{\mathrm{Na}} = 4.129$ and
$\hat\beta_{\mathrm{Li}} = 4.137$ (both $\times10^{-15}$\,V$\cdot$s), differing
by 0.19\%.  A pooled estimate weighting by $S_{\nu\nu,m}$ gives
$\hat\beta_{\mathrm{pool}} \approx 4.133\times10^{-15}$\,V$\cdot$s,
with a standard error reduced by roughly $1/\sqrt{2}$ relative to either
single-metal estimate.

The Wald statistic for testing the slope-equality restriction
$H_0\colon\beta_{\mathrm{Na}} = \beta_{\mathrm{Li}}$ is
\[
  z = \frac{\hat\beta_{\mathrm{Na}} - \hat\beta_{\mathrm{Li}}}
      {\sqrt{\mathrm{SE}^2(\hat\beta_{\mathrm{Na}})
           + \mathrm{SE}^2(\hat\beta_{\mathrm{Li}})}}
  \approx \frac{0.008\times10^{-15}}{2\times4.6\times10^{-18}}
  \approx 0.87.
\]
The test statistic is well below the critical value of 1.96, so the
data provide no evidence against universality of $h/e$ across sodium
and lithium --- a confirmation of a key structural prediction of the
quantum hypothesis.

%======================================================================
\section{The Contact-EMF Identification Problem}
\label{sec:contact}

The contact EMF $v_c$ between the photoemitting metal and the collector is
a classic \emph{nuisance parameter}.  In the language of regression,
equation~\eqref{eq:reg} has an intercept $\alpha = -W/e + v_c$ that
conflates physics ($W/e$, the work function) with apparatus ($v_c$).

\paragraph{Pre-Millikan experiments.}  The early experiments did not measure
or control $v_c$.  If $v_c$ drifted between measurements at different
frequencies, the apparent stopping potentials would shift inconsistently,
biasing \emph{both} slope and intercept.  Hughes and Richardson \& Compton
measured different lines on different days with no guarantee of constant $v_c$,
which Millikan \citeyearpar{Millikan1916} identified as a contributing cause
of their $h$ values being 10--45\% below Planck's.

\paragraph{Millikan's solution.}  Millikan measured all six spectral lines
in a single continuous session after a freshly shaved surface had stabilised
(typically 1--2 hours after shaving).  He verified that $v_c$ remained constant
to $\pm 0.01$\,V over the session by using the Kelvin double-balance.  When
$v_c$ is constant across all observations, it enters as a pure intercept shift
and does not affect $\hat\beta$.  The slope $\hat\beta = h/e$ is then estimated
without bias regardless of the unknown value of $v_c$.

\paragraph{Separate identification of $W$.}  By independently measuring $v_c =
2.51$\,V (Kelvin method) and reading the intercept $\hat\alpha = -1.81$\,V
from the regression (equation~\eqref{eq:alphahat_num}), Millikan recovered
\begin{align*}
  W_{\mathrm{Na}}/e &= v_c - \hat\alpha = 2.51 - (-1.81) = 4.32\text{ eV},\\
  W_{\mathrm{Na}} &= 4.32\times1.602\times10^{-19} = 6.92\times10^{-19}\text{ J}.
\end{align*}
However, he noted that the exact work function of a freshly scraped sodium
surface in $10^{-2}$\,mmHg residual gas was not well characterised.  The
widely quoted value for sodium is $W_{\mathrm{Na}} = 2.36$\,eV (modern value
for clean bulk Na).  The larger intercept-implied value reflects the unusual
surface conditions (freshly shaved Na in partial vacuum with absorbed
residual gas).

%======================================================================
\section{Nine Two-Point Determinations}
\label{sec:twopoint}

In addition to the direct regression, Millikan independently estimated the
slope from nine pairs of spectral lines, always pairing one line with 5461\,\AA\
(chosen for its high intensity and hence most precise balance point).

\begin{table}[ht]
\centering
\caption{Millikan (1916), Table~II: nine two-point slope determinations
for sodium.  Each row is an independent measurement session with a fresh
surface shaving.  The mean and standard deviation are computed from these
nine observations.}
\label{tab:two_point}
\begin{tabular}{ccc}
\toprule
Session & $\lambda$ paired with 5461\,\AA & $\hat\beta$ ($10^{-15}$\,V$\cdot$s) \\
\midrule
1 & 3126\,\AA & 4.11 \\
2 & 3650\,\AA & 4.14 \\
3 & 3126\,\AA & 4.10 \\
4 & 3650\,\AA & 4.12 \\
5 & 3126\,\AA & 4.24 \\
6 & 4047\,\AA & 3.98 \\
7 & 2535\,\AA & 4.04 \\
8 & 3126\,\AA & 4.24 \\
9 & 4047\,\AA & 4.21 \\
\midrule
Mean   & & $\mathbf{4.131}$ \\
Std.~dev.\ ($n=9$) & & $0.087$ \\
Std.~error of mean & & $0.029$ \\
95\% CI & & $[4.06,\; 4.20]$ \\
\bottomrule
\end{tabular}
\end{table}

\noindent The two-point mean $4.131\times10^{-15}$\,V$\cdot$s agrees with
the five-point OLS estimate $4.129\times10^{-15}$\,V$\cdot$s to better than
0.1\%.  The variation across sessions (sessions 5 and 8 giving 4.24 vs.\
sessions 3 and 6 giving 3.98--4.10) reflects day-to-day instrumental noise
and small variations in surface condition, consistent with Millikan's stated
per-measurement uncertainty of 1--4\%.

%======================================================================
\section{Comparison of Historical Estimates}
\label{sec:comparison}

Table~\ref{tab:comparison} compares the estimates of $h$ from all major
pre-1920 photoelectric experiments.

\begin{table}[ht]
\centering
\caption{Historical estimates of Planck's constant $h$ from photoelectric
experiments.  Millikan's attribution of earlier errors is discussed in
\citet[pp.\ 373--375]{Millikan1916}.  ``Error'' is relative to the modern
NIST CODATA 2022 value $h = 6.6261\times10^{-27}$\,erg$\cdot$s.}
\label{tab:comparison}
\begin{tabular}{p{4.2cm}rcrc}
\toprule
Authors & Year & Metal(s) & $h$ ($10^{-27}$\,erg$\cdot$s) & Error \\
\midrule
Hughes        & 1912--13 & various & 3.55--5.85 & $-$12 to $-$46\% \\
Richardson \& Compton & 1912 & 8 metals & 3.5--5.8 & $-$12 to $-$47\% \\
Millikan (Na, 5-pt)  & 1916 & Na & 6.573 & $-$0.80\% \\
Millikan (Na, 9-pt)  & 1916 & Na & 6.578 & $-$0.73\% \\
Millikan (Li)        & 1916 & Li & 6.584 & $-$0.64\% \\
Millikan (Na+Li mean)& 1916 & Na+Li & \textbf{6.579} & $\mathbf{-0.72\%}$ \\
Planck (black body)  & 1913 & --- & 6.415 & $-$3.19\% \\
\midrule
Modern NIST 2022     & 2022 & --- & 6.6261 & 0 \\
\bottomrule
\end{tabular}
\end{table}

The pattern is striking: all pre-Millikan estimates are biased downward by
10--47\%, with a bias attributable entirely to stray short-wavelength light
(raises apparent $V_0$ at long-wavelength lines, flattening the slope),
restricted frequency range (enlarges SE of slope), and uncontrolled contact
EMF (adds noise and possible trend bias to the intercept).  Millikan's
engineering solutions---fresh surfaces, filtered lines, wide frequency range,
Kelvin $v_c$ measurement---removed all three sources of bias simultaneously.

%======================================================================
\section{Millikan's Epistemic Reluctance}
\label{sec:millikan_reluctance}

Perhaps the most instructive aspect of this episode for a statistician is
the gap between \emph{confirming an equation} and \emph{accepting a model}.

Millikan's 1916 paper confirmed Einstein's equation~\eqref{eq:einstein} to
within 0.5\%.  Yet Millikan wrote:
\begin{quote}
  ``Despite then the apparently complete success of the Einstein equation,
  the physical theory of which it was designed to be the symbolic expression
  is found so untenable that Einstein himself, I believe, no longer holds
  to it.'' \citep[p.\ 388]{Millikan1916}
\end{quote}
The light-quantum hypothesis was, in Millikan's judgment, physically
implausible because it conflicted with the wave theory of diffraction and
interference.  He had hoped his experiment would \emph{refute} the equation,
and was surprised when it confirmed it.

By 1923---after Compton scattering \citep{Compton1923} provided independent
evidence for the particulate nature of light---Millikan conceded:
\begin{quote}
  ``This work resulted, contrary to my own expectation, in the first direct
  experimental proof \ldots of the complete validity of the Einstein equation.''
  \quad (Nobel address, 1923; quoted by \citealt{Holton1999})
\end{quote}

The lesson for statistical inference: a sufficiently precise measurement of
a regression coefficient can confirm a \emph{specific functional form} while
leaving the \emph{causal or structural interpretation} contested.  Millikan
confirmed the slope $h/e$ to 0.5\% without accepting the physical model
that gives it meaning.  This is analogous to a macroeconomist confirming a
reduced-form predictive relationship without endorsing the structural DSGE
model that motivated it.

%======================================================================
\section{Einstein as Machine Learner}
\label{sec:ml}

\subsection{Overview}

The vocabulary of modern machine learning---supervised learning, hypothesis
class, loss function, inductive bias, feature engineering, transfer
learning---did not exist in 1905.  Yet the structure of Einstein's
scientific reasoning maps onto it with striking precision.  This section
makes the correspondence explicit.  The goal is not to reduce Einstein's
achievement to a computational exercise, but to show that the epistemological
operations he performed are precisely the ones formalised in the modern theory
of statistical learning --- and to show where that analogy has limits.

\subsection{Supervised Learning and Labeled Data}

The photoelectric dataset is a standard supervised regression problem.
The \emph{input feature} is the frequency~$\nu$ of the incident light;
the \emph{label} is the stopping potential~$V_0$.  The task is to learn
a function $f:\nu\mapsto V_0$ that generalises to unobserved frequencies.

Lenard (1902) provided the \emph{qualitative} version: the label was observed
(high $V_0$ at high frequency) but not quantified reliably.  Millikan (1916)
provided the \emph{quantitative} training set: six labeled pairs
$(\nu_i, V_{0i})$ for each metal, with measurement precision sufficient to
estimate the slope $h/e$ to 0.5\%.

\subsection{The Hypothesis Class and Inductive Bias}
\label{sec:hyp_class}

In statistical learning theory, a \emph{hypothesis class} $\mathcal{H}$ is the
set of functions the learner considers.  The choice of $\mathcal{H}$ encodes
the learner's \emph{inductive bias}: prior knowledge about the structure of
the true function.

Einstein's hypothesis class is:
\begin{equation}\label{eq:hyp_class}
  \mathcal{H} = \bigl\{f_{\alpha,\beta}:\nu\mapsto\alpha+\beta\nu
  \bigm|\alpha\in\mathbb{R},\;\beta>0\bigr\},
\end{equation}
together with the \emph{cross-metal constraint} that $\beta$ is the same
for all metals.  This is an extremely tight class: a one-parameter family
of parallel lines.

The restriction to $\mathcal{H}$ was not ad hoc.  It followed deductively
from the quantum hypothesis $E=h\nu$ and energy conservation for photoemission.
The quantum hypothesis \emph{is} the inductive bias: it restricts the
hypothesis class from all possible functions of $(\nu, I, \lambda, \ldots)$
to a specific affine form in $\nu$ alone.  Without this bias, five data points
are hopelessly insufficient --- infinitely many smooth functions pass through
five points.  The quantum hypothesis provides the Occam's-razor constraint
that makes learning possible with minimal data.

\subsection{Feature Engineering: Choosing $\nu$ Over Intensity
and Wavelength}

A critical step before any learning occurs is choosing the \emph{input features}.
Lenard's experiments implicitly performed a feature importance analysis.  He
measured $V_0$ as a function of both frequency (colour) and intensity~$I$.  His
result was unambiguous: $V_0$ varies strongly with $\nu$ but not at all with~$I$.
Intensity had zero feature importance; $\nu$ had all the importance.

Classical wave theory predicted the opposite: wave amplitude $A \propto
\sqrt{I}$ should determine energy delivered to an electron, so $V_0$ should
grow with $I$.  The empirical irrelevance of $I$ was the anomaly that
motivated Einstein's model.  In modern ML language, Lenard performed an
experiment that eliminated $I$ from the feature set and selected $\nu$---the
correct and sufficient input feature---before any parametric fitting was done.

\subsection{The Loss Function and Training}

Once the hypothesis class~\eqref{eq:hyp_class} and the feature $\nu$ are
fixed, the learning objective is to choose $(\alpha,\beta)$ to minimise the
empirical risk under squared-error loss:
\[
  \hat{R}(\alpha,\beta) = \frac{1}{n}\sum_{i=1}^n(V_{0i} - \alpha - \beta\nu_i)^2.
\]
This is identical to the MSE loss used in neural-network regression.  The
minimiser of $\hat{R}$ is the OLS/MLE estimator derived in
Sections~\ref{sec:regression}--\ref{sec:mle}: training is the closed-form
solution $\hat\beta = S_{\nu V}/S_{\nu\nu}$.  For Einstein's linear model,
the loss surface is a convex paraboloid in $(\alpha,\beta)$ with a unique
global minimum; no stochastic gradient descent is needed.

\subsection{Generalisation and Transfer Learning}

\paragraph{Generalisation.}
The fitted line $\hat V_0 = \hat\alpha + \hat\beta\nu$ predicts stopping
potentials for frequencies \emph{not in the training set}.  The near-zero
residuals (Section~\ref{sec:regression}) show no overfitting.  In
learning-theoretic terms, the Vapnik-Chervonenkis (VC) dimension of
$\mathcal{H}$ (a one-dimensional affine class) equals~2, so five training
points are more than sufficient for reliable generalisation with high
probability.

\paragraph{Transfer learning.}
A model trained on sodium data successfully predicts lithium stopping
potentials up to an intercept shift.  The slope $\hat\beta = h/e$ is
\emph{transferred exactly} from one metal to another: it is a universal
representation.  In modern parlance, training on sodium provides a
pre-trained feature extractor (the slope); the lithium task requires only
fine-tuning the intercept $\alpha_{\mathrm{Li}}$, which is the metal-specific
work-function parameter.  The Wald test of
Section~\ref{sec:panel} confirms that the zero-shot transfer of the slope
from Na to Li fails to be rejected by the data ($z = 0.87$).

\subsection{Model Selection: Quantum vs.\ Classical Wave Theory}

At the time of Einstein's paper two models were in competition:
\begin{itemize}
  \item \textbf{Classical wave model.}  Intensity determines energy transfer;
    $V_0$ increases with wave amplitude $A \propto \sqrt{I}$.  This predicts
    zero $\nu$-dependence in~$V_0$, i.e., $R^2 = 0$ on the $(\nu, V_0)$ data.
  \item \textbf{Quantum model.}  $V_0 = (h/e)\nu - W/e$.  This predicts
    an exact linear $\nu$-dependence with a specific slope that is universal
    across metals.
\end{itemize}
Millikan's data delivers $R^2 \approx 0.99999$ for the quantum model.
Any modern model selection criterion --- AIC, BIC, cross-validation, Bayes
factor --- would produce essentially infinite evidence in favour of the
quantum model over the classical alternative.

\subsection{What a Naive Machine Learning System Would Miss}

It is instructive to ask what a purely data-driven ML system, given only
the table of $(\nu_i, V_{0i})$ pairs, would learn.

\begin{enumerate}
  \item A linear regression package would correctly recover
    $\hat\beta = 4.129\times10^{-15}$\,V$\cdot$s and $R^2 \approx 1$.
  \item It would \emph{not} know that $\hat\beta$ identifies a fundamental
    constant of nature.  The number $4.129\times10^{-15}$ is physically
    meaningless without the quantum hypothesis that gives it the
    interpretation $h/e$.
  \item It would \emph{not} impose the cross-metal universality
    restriction $\beta_{\mathrm{Na}} = \beta_{\mathrm{Li}}$ from first
    principles.  The universality is a prediction of the quantum hypothesis;
    from a purely data-driven perspective, it is just a coincidence to be
    discovered, not a constraint to be imposed.
  \item Modern symbolic regression systems such as AI Feynman
    \citep{Udrescu2020} can, in principle, rediscover the linear form from
    data.  But they require knowledge of units and physical constants to
    identify $\hat\beta$ with $h/e$ rather than reporting it as an
    unexplained numerical coefficient.
\end{enumerate}

Einstein's irreplaceable contribution was precisely where modern ML is
weakest: supplying the theoretical framework that gives regression
coefficients their physical meaning.  Specifying the hypothesis class required
more intellectual effort than fitting the line.

\subsection{The Deductive--Inductive Loop}

Modern discussions of ``AI for science'' often describe a purely inductive
paradigm: given enough data, learn the law.  The photoelectric episode
illustrates a more powerful paradigm that has driven physical science since
Newton: the \emph{hypothetico-deductive loop}.

\begin{table}[ht]
\centering
\caption{The hypothetico-deductive loop in the photoelectric episode,
translated into the language of modern machine learning.}
\label{tab:loop}
\begin{tabular}{p{4cm}p{4cm}p{4cm}}
\toprule
Step & Scientific name & ML name \\
\midrule
Propose $E = h\nu$ & Theoretical hypothesis & Model architecture \\
Derive $V_0 = (h/e)\nu - W/e$ & Deduction &
  Hypothesis class \\
Choose $\nu$ not $I$ & Feature selection & Feature engineering \\
Measure $(\nu_i,V_{0i})$ & Experiment & Training data \\
Min.\ $\sum(V_{0i}-\hat V_{0i})^2$ & Least squares &
  MSE training \\
$|\hat\beta - h/e| < 1\%$ & Confirmation &
  Test accuracy \\
Na and Li: same $\hat\beta$ & Universality & Zero-shot transfer \\
\bottomrule
\end{tabular}
\end{table}

The deductive step is the engine of efficiency: because the physics
prescribes the functional form, the number of free parameters ($\alpha, \beta$),
and even the units of the slope, five training points are sufficient to
estimate $h$ to 0.1\%.  A purely inductive system that must discover the functional
form from data faces a drastically larger sample-complexity problem.

This is both the power and the limit of current AI for scientific discovery.
Systems such as AI Feynman or neural-network-based symbolic regression can
\emph{in principle} induce the linear law from data, but they do so by
exhaustive search over a combinatorially large space of functional forms,
requiring far more data and computation than Einstein's one-paragraph
derivation.  The human contribution --- theory construction --- remains
indispensable.

%======================================================================
\section{A Deep Neural Network on the Photoelectric Data}
\label{sec:dnn}

\subsection{Two Hypothesis Classes Compared}

Einstein's model and a deep neural network (DNN) represent opposite
extremes of the bias-variance spectrum for the photoelectric regression.
Einstein's hypothesis class is equation~\eqref{eq:hyp_class}: a
one-parameter family of parallel lines with VC dimension equal to~2.
The Gauss--Markov theorem and the Cram\'er--Rao bound together guarantee
that OLS achieves minimum variance among all unbiased estimators;
five training points are more than sufficient.

A DNN with $L$ hidden layers and $N$ neurons per layer is a universal
function approximator: for any continuous target function on a compact
interval, there exists a network that approximates it to arbitrary
precision \citep{Hornik1989}.  For a one-dimensional input ($\nu$) and
one-dimensional output ($V_0$), even a modest architecture---three
hidden layers with 16 ReLU neurons each---has approximately
\[
  p = (1\cdot16+16)+(16\cdot16+16)+(16\cdot16+16)+(16\cdot1+1)
    = 593
\]
free parameters.  The VC dimension of this network is of order $p\log p
\approx 5{,}700$ \citep{Vapnik1998}, meaning that reliable generalisation
to unseen frequencies requires on the order of $n\gtrsim 5{,}700$
labeled training examples for 1\% accuracy with high probability.  The
five points Millikan assembled in 1916 fall short by three orders of
magnitude.

\subsection{Epoch I---Lenard's Qualitative Findings (1902)}

Lenard established two qualitative facts: stopping potential rises with
frequency and is independent of intensity.  There are no numerical
$(\nu_i,\,V_{0i})$ pairs.  A DNN requires labeled numerical examples;
with none, it cannot begin training.  Einstein required zero labeled
examples: the quantum hypothesis plus energy conservation yielded
equation~\eqref{eq:einstein} as a deduction.  The asymmetry is stark---
Einstein's ``model fitting'' was a one-paragraph derivation;
the DNN's fitting cannot start.

\subsection{Epoch II---Hughes and Richardson \& Compton (1907--1914)}

The intermediate experiments produced roughly 25--40 numerical
$(\nu_i,\,V_{0i})$ pairs, clustered in a narrow 60\,nm UV range with
uncontrolled contact EMFs and stray scattered light.
Three problems foreclose reliable DNN learning.

\paragraph{Underdetermination.}
With $n \approx 30$ and $p = 593$, the DNN is massively overparameterised.
Gradient descent finds a minimum-norm interpolant that passes through
all 30 training points exactly (zero training loss), but is effectively
unconstrained between them.  Applying $\ell_2$ weight decay suppresses
oscillations, but the optimal regularisation strength cannot be selected
without a held-out validation set; with $n=30$ this is statistically
isolated.

\paragraph{Restricted frequency range.}
Over a 60\,nm UV band ($\Delta\nu \approx 0.7\times10^{14}$\,Hz), any
smooth function is locally indistinguishable from a polynomial of any
degree.  The DNN has no mechanism to infer that the global functional
form is linear rather than quadratic or square-root; all three fit
equally well on so narrow an interval.

\paragraph{Systematic label contamination.}
Uncontrolled contact EMFs imply that the effective label for observation
$i$ is $V_{0i} + \delta_i$, where $\delta_i$ is unknown and
observation-specific.  The DNN receives contaminated targets and,
lacking any physical model of the contaminant, cannot separate the
linear physical trend from the systematic noise.  The learned function
absorbs both, producing a biased and uninterpretable output.

\subsection{Epoch III---Millikan's Data (1916)}

Millikan's five-point sodium dataset is the cleanest photoelectric
dataset of the era.  What does a DNN find?

\paragraph{The interpolation regime.}
With $n=5$ and any DNN with $p>5$ (which includes every
useful architecture), the network operates in the interpolation
regime: it memorises all five training points exactly,
producing zero training loss \citep{Belkin2019}.  The resulting
function is unconstrained everywhere between the training points.
Even the best-case minimum-norm interpolant---``well-behaved'' in the
double-descent sense---provides no basis for slope inference: there
are no residual degrees of freedom with which to estimate the noise
variance or a standard error on the slope.

\paragraph{Loss of inferential machinery.}
OLS with $n=5$ and $p=2$ free parameters uses $n-2=3$ residual
degrees of freedom to estimate $\hat\sigma^2$ and hence
$\mathrm{SE}(\hat\beta)$.  Any DNN with more than three parameters
exhausts those degrees of freedom entirely.  The DNN cannot report
a confidence interval for $h/e$, cannot test whether the slope
differs from zero, and cannot quantify the 0.1\% relative precision
that OLS achieves at the Cram\'er--Rao bound.

\paragraph{No basis for universality.}
The universality of the slope $\beta = h/e$ across metals is a
theoretical prediction of the quantum hypothesis, not an empirical
feature discoverable from sodium data alone.  A DNN trained on
sodium would need to be retrained from scratch on lithium.  The
only mechanism by which the slope could transfer automatically from
Na to Li is an explicit architectural constraint that encodes the
universality---which is equivalent to importing the physics into
the model.

\subsection{The Modern Era: Large Datasets}

By the mid-twentieth century, photoelectric measurements spanned
decades of metals, surface preparations, and photon energies.
Pooling even a modest subset---say $n=5{,}000$ clean observations
over two octaves of frequency for a single metal---places a DNN
firmly in the classical regime ($n\gg p$).

With smooth activations and appropriate regularisation, the
DNN would converge, as $n\to\infty$, to the
minimum-population-risk function in its hypothesis class.  Because
the true function is exactly linear, the minimum-risk function in
any sufficiently rich class \emph{is} the linear function.  The
DNN would eventually recover the line.  What it still could not do:
\begin{itemize}
  \item Identify the learned slope $\hat{w} \approx
    4.129\times10^{-15}$\,V$\cdot$s as the ratio $h/e$ of two
    independently measured constants.  The slope appears as a
    numerical weight, not as a ratio of fundamental quantities.
  \item Impose the cross-metal constraint $\beta_{\mathrm{Na}} =
    \beta_{\mathrm{Li}} = \cdots$ from first principles.  Universality
    would appear as an empirical pattern to be confirmed, not a
    structural prior to be imposed.
  \item Decompose the intercept into a physical work function $W/e$
    and an apparatus nuisance $v_c$ without independent
    measurements of $v_c$.
  \item Reduce the required sample from $O(10^3)$ examples to five.
    A correctly specified two-parameter linear model needs five points;
    no architectural innovation can replicate this compression without
    importing the physics.
\end{itemize}

Modern symbolic regression systems such as AI Feynman \citep{Udrescu2020}
can, in principle, rediscover the linear form and even match $\hat{w}$ to
$h/e$ if given units and physical constants as hardcoded inputs.  But
that hardcoding is itself a form of prior physical knowledge---the quantum
hypothesis in disguise---not something extracted from the $(\nu,V_0)$
data.

\subsection{Summary: Sample Complexity and the Value of Theory}

Table~\ref{tab:sample_complexity} quantifies the contrast.

\begin{table}[ht]
\centering
\caption{Sample-complexity comparison between Einstein's two-parameter
linear model (VC dimension $= 2$, derived from the quantum hypothesis) and
a three-layer DNN with 16 neurons per layer ($p=593$ parameters,
VC dim $\approx 5{,}700$).  The ``sufficient $n$'' for a 1\% generalisation
error at 95\% confidence is $O(\text{VC dim}/\epsilon^2)$.}
\label{tab:sample_complexity}
\begin{tabular}{p{2.8cm}rp{2.4cm}p{2.4cm}p{4.0cm}}
\toprule
Epoch & $n$ available & Einstein model & DNN (VC$\approx$5700) & DNN outcome \\
\midrule
Lenard 1902   & $\approx 0$  & Sufficient$^\dagger$ & Insufficient
  & Cannot train (no labels) \\
Hughes 1912   & $\sim 30$    & Sufficient
  & Insufficient ($\times$190) & Memorises; no generalisation \\
Millikan 1916 & 5            & Sufficient
  & Catastrophically insufficient & Interpolates; no inference \\
Modern era    & $\sim 5000$  & Sufficient
  & Sufficient & Recovers linear shape;
  cannot identify $h/e$ \\
\bottomrule
\multicolumn{5}{p{14cm}}{$^\dagger$ The functional form is derived from
  theory; only two free parameters remain to be estimated from data.}
\end{tabular}
\end{table}

The lesson is not that DNNs are poor learners in general.  For
high-dimensional problems where no parsimonious parametric family is
known---image recognition, protein structure prediction, language
modelling---the DNN's universal-approximation capacity is an enormous
advantage.  The photoelectric problem sits at the opposite extreme: a
one-dimensional input, a one-dimensional output, and five to thirty
training examples.  In this setting the DNN's flexibility is a liability,
not an asset.

Einstein's theoretical act---proposing $E=h\nu$---reduced the effective
hypothesis space from the infinite-dimensional function class of a DNN to
a two-dimensional manifold of parallel lines.  This compression reduced
the required number of training examples by three orders of magnitude.
No architectural innovation or training algorithm can replicate that
compression without importing the physics; it was achieved entirely in the
mind of a 26-year-old patent clerk, before a single labeled example was
available.

%======================================================================
\section{Conclusions}
\label{sec:conclusions}

Einstein's photoelectric equation is a single-regressor linear model with
one slope ($h/e$) and one intercept ($-W/e + v_c$).  The central estimation
challenge is not the regression itself---which is trivial given clean data---but
obtaining clean data: fresh metal surfaces free of oxide (eliminates
work-function noise), high vacuum (eliminates atmospheric contamination),
spectrally pure light (eliminates bias from stray UV), a wide frequency range
(reduces SE of slope), and control of the nuisance parameter $v_c$ (enables
identification of both slope and intercept).

Millikan's decade-long experimental program was, in essence, a campaign to
raise data quality until the OLS standard errors were small enough to
distinguish a specific value of the slope ($h/e = 4.136\times10^{-15}$\,V$\cdot$s)
from the alternatives implied by earlier, discrepant theories.

The final five-point sodium regression gives
\begin{align*}
  \hat\beta &= 4.129\times10^{-15}\text{ V$\cdot$s}
             \quad (h = 6.61\times10^{-34}\text{ J$\cdot$s}, \text{ 0.5\% above modern}), \\
  \hat\alpha &= -1.813\text{ V}
             \quad (W_{\mathrm{Na}}/e \approx 4.3\text{ eV given } v_c=2.51\text{ V}), \\
  R^2 &\approx 0.99999.
\end{align*}

The episode illustrates three principles that recur across scientific
disciplines: (i) theoretical priors guide experimental design by specifying
the functional form to be estimated; (ii) data quality---not statistical
sophistication---is the binding constraint for precision; and (iii) confirming
a regression equation is not the same as confirming the theoretical model
that produced it.

%======================================================================
\appendix

\section{Einstein's Equation in Modern Physics}
\label{app:modern}

\subsection{The Equation Remains Exact}

Einstein's photoelectric equation~\eqref{eq:einstein} is not an
approximation to a deeper expression---it is an exact consequence of
quantum electrodynamics (QED).  In QED the photoelectric effect is
described as the absorption of a single photon by an electron near
the Fermi level of a metal; energy conservation yields
equation~\eqref{eq:einstein} directly.  The work function $W$ is the
electrochemical potential required to remove an electron from the Fermi
level to the vacuum level; modern density-functional theory computes $W$
to within a few percent for clean surfaces.  Millikan's regression
therefore tests an exact result of QED, not an empirical approximation.

The equation is exploited continuously in surface science.  Ultraviolet
Photoelectron Spectroscopy (UPS) and Kelvin probe instruments apply
equation~\eqref{eq:einstein} to measure work functions
of semiconductor films, organic photovoltaic layers, and heterogeneous
catalysts with sub-meV resolution.

\subsection{X-ray Photoelectron Spectroscopy}
\label{app:xps}

When the incident photon energy $h\nu$ is drawn from the X-ray range
(100\,eV--10\,keV), the emitted electron's kinetic energy $eV_0$
identifies the \emph{binding energy} $E_B$ of the core electron level
from which it was ejected \citep{Siegbahn1967}:
\begin{equation}\label{eq:xps}
  E_B = h\nu - eV_0 - \phi_{\mathrm{spec}},
\end{equation}
where $\phi_{\mathrm{spec}}$ is the spectrometer work function, a
calibration constant.  Equation~\eqref{eq:xps} is
equation~\eqref{eq:einstein} rearranged: the unknown is now the binding
energy rather than $h/e$.  Because each element has a unique set of
core-level binding energies, X-ray Photoelectron Spectroscopy (XPS/ESCA)
provides quantitative, element-specific chemical analysis of surfaces.
Kai Siegbahn received the 1981 Nobel Prize in Physics for this technique,
which is now one of the most widely used methods in materials science,
semiconductor device fabrication, and surface catalysis.

In statistical terms, XPS is a multivariate extension of Millikan's
regression: dozens of binding energies are estimated simultaneously by
inverting Einstein's linear equation at multiple photon energies.

\subsection{Multi-Photon Photoelectric Effect}
\label{app:multiphoton}

Einstein's equation assumes each emitted electron absorbs exactly one
photon.  At the peak intensities of pulsed lasers
($I \gtrsim 10^{10}$\,W/cm$^2$), an electron can sequentially absorb
$n > 1$ photons before escaping the surface.  The $n$-photon
generalisation of equation~\eqref{eq:einstein} is
\begin{equation}\label{eq:multiphoton}
  eV_0 = n\,h\nu - W, \qquad n = 1, 2, 3, \ldots
\end{equation}
Above-threshold ionisation (ATI)---photoelectron emission at kinetic
energies exceeding the single-photon threshold by integer multiples of
$h\nu$---was first observed by \citet{Agostini1979}.  The resulting
photoelectron spectrum displays a comb of peaks separated by $h\nu$,
each satisfying equation~\eqref{eq:multiphoton} for the corresponding
$n$.  In the language of Section~\ref{sec:ml}, ATI extends Einstein's
hypothesis class to allow non-unit photon multiplicity; the slope remains
$h/e$ and the universal structure is preserved across metals.

\subsection{Does the Photoelectric Effect Prove the Quantisation of Light?}
\label{app:lamb_scully}

Section~\ref{sec:millikan_reluctance} noted Millikan's epistemic reluctance
to accept the light-quantum hypothesis despite confirming its equation.
Modern physics provides a more precise statement of what Millikan's
regression actually proves.

\citet{Lamb1969} showed that the photoelectric stopping-potential equation
can be derived from a \emph{semiclassical model}: the electromagnetic
field is treated as a classical wave while only the atom (electron) is
quantised.  In this model a classical wave drives transitions between
quantised atomic energy levels, and the ejected-electron kinetic energy
still satisfies equation~\eqref{eq:einstein}.  The reason is that
equation~\eqref{eq:einstein} follows from energy conservation and the
quantised \emph{atomic} level structure alone; it does not require the
radiation field itself to be quantised.

This observation does not diminish Einstein's achievement.  He identified
the correct hypothesis class~\eqref{eq:hyp_class} and the correct
functional form, and the slope $h/e$ encodes Planck's constant whether
or not the field is ``really'' composed of quanta.  But the
Lamb--Scully result clarifies what Millikan's regression \emph{proves}:
it confirms the functional form $V_0 \propto \nu$ and measures $h/e$
to 0.5\%, and it is consistent with the photon hypothesis---but it does
not distinguish the quantised-field hypothesis from semiclassical
alternatives.  Millikan's epistemic reluctance was, ironically, more
defensible than the standard narrative suggests.

\subsection{Photon Antibunching: Definitive Evidence for Field Quantisation}
\label{app:antibunching}

The genuine experimental proof that the radiation field must be quantised
came not from the photoelectric effect but from photon-counting
experiments in quantum optics.  The relevant observable is the
statistical regularity---or irregularity---with which successive photons
are emitted.

\citet{HanburyBrown1956} observed \emph{photon bunching} in thermal
(chaotic) light: photon pairs from a thermal source arrive at two
detectors with a coincidence rate exceeding a Poisson benchmark, reflecting
the Bose-Einstein statistics of thermal photons.  \citet{Glauber1963}
developed the quantum theory of optical coherence, introducing coherent
states of the radiation field (the states most closely resembling classical
waves) and deriving the photon-number statistics that separate classical,
coherent, and thermal light.  Glauber received the 2005 Nobel Prize in
Physics for this framework.

The decisive evidence for field quantisation is photon
\emph{antibunching}: the observation that successive photons from a
single quantum emitter are \emph{less} likely to arrive together than
Poisson statistics predict.  \citet{Kimble1977} first observed
sub-Poissonian photon-number statistics in the resonance fluorescence of
individual sodium atoms.  Antibunching is impossible in any classical or
semiclassical model of the electromagnetic field: a classical field,
however modulated or noisy, always produces at minimum Poissonian
coincidences.  The antibunching experiment is therefore the decisive
model-selection test---in the language of Section~\ref{sec:ml}---that
rules out the semiclassical hypothesis of \citeauthor{Lamb1969} and
establishes that the radiation field must be quantised.

The historical sequence is instructive.  Einstein's 1905 paper introduced
the photon as a \emph{heuristic}---his own word in the title---and
Millikan confirmed its equation to 0.5\% in 1916.  Definitive experimental
evidence that the radiation field is itself quantised arrived only in 1977,
by a route---photon-number statistics in resonance fluorescence---that
Einstein's equation neither anticipated nor directly implied.  The
regression confirmed the equation; quantum optics experiments seventy
years later confirmed the physical interpretation.

\subsection{Implications for the Machine Learning Reinterpretation}
\label{app:ml_implications}

The modern results above sharpen several points from Section~\ref{sec:ml}.

\paragraph{Depth of the hypothesis class.}
Einstein chose hypothesis class~\eqref{eq:hyp_class} by a heuristic
argument whose full justification---QED, Fermi's golden rule, the
Fermi-Dirac distribution of electrons in a metal---emerged over the
following three decades.  The class proved correct for reasons deeper
than those available in 1905, and it extends over a five-order-of-magnitude
range of photon energies (visible UV to hard X-rays) without modification.
The ``inductive bias'' encoded in the two-parameter linear model is even
more powerful than Section~\ref{sec:ml} indicates.

\paragraph{What the regression tests.}
Millikan's regression tests the \emph{functional form} $V_0 \propto \nu$
and estimates the slope $h/e$.  The Lamb--Scully result shows that this
functional form and this slope are consistent with both quantised-field
and semiclassical models of radiation.  Distinguishing these model
classes requires a different observation---photon-number variance---and a
different experimental design.  In the language of Section~\ref{sec:ml}:
the regression identifies the best-fitting model \emph{within} a given
hypothesis class, but selecting \emph{between} hypothesis classes (quantised
vs.\ semiclassical field) requires a different objective function and a
different experiment.

\paragraph{Theory specificity in experimental design.}
The antibunching experiments of \citet{Kimble1977} were designed
specifically to test the quantised-field hypothesis; the relevant
observable (photon-number variance) follows directly from
Glauber's coherence theory.  Without the quantised-field hypothesis and
its prediction of sub-Poissonian statistics, there was no reason to
measure photon-number correlations at all.  This repeats the lesson of
Millikan's experiment: theory specifies the hypothesis class and, from
it, the decisive test.  The progression from Millikan (1916) to
\citeauthor{Kimble1977} (1977) illustrates how each theoretical
refinement generates a new, more discriminating experimental test---a
chain of hypothetico-deductive loops (Table~\ref{tab:loop}) whose links
are forged by theory, not by data alone.

\bigskip
\bibliographystyle{plainnat}
\bibliography{einstein_photoelectric_ML}

\end{document}
