% riess_schmidt_perlmutter_ML.tex
% "The Accelerating Universe as Statistical Inference:
%  Type Ia Supernovae, Maximum Likelihood, and the
%  Cosmological Constant, 1988--1999"
% Thomas J. Sargent, 2026

\documentclass[12pt]{article}

\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage[margin=1.2in]{geometry}
\usepackage{setspace}
\usepackage{microtype}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{parskip}

\onehalfspacing

\hypersetup{
  colorlinks  = true,
  linkcolor   = blue,
  citecolor   = blue,
  urlcolor    = blue
}

\title{The Accelerating Universe as Statistical Inference:\\
       Type Ia Supernovae, Maximum Likelihood,\\
       and the Cosmological Constant, 1988--1999}
\author{Thomas J.\ Sargent}
\date{April 2026}

\begin{document}
\maketitle

\begin{abstract}
In their Nobel Prize--winning papers, Adam Riess, Brian Schmidt, and their
High-Z Supernova Search Team \citep{Riess1998}, and independently Saul
Perlmutter and the Supernova Cosmology Project \citep{Perlmutter1999},
reported that the expansion of the universe is \emph{accelerating}.
Their argument had two components: a physical \emph{model} (the
Friedmann equations with a cosmological constant $\Lambda$) and a
\emph{measurement equation} relating the observable apparent magnitude
of a Type Ia supernova to its redshift, with the shape of that relation
controlled by two density parameters~$\Omega_M$ and~$\Omega_\Lambda$.
This note examines their reasoning from a modern statistical perspective.
I describe the data assembled by both groups---the Cal\'an/Tololo
low-redshift training sample and the high-redshift supernova discoveries
at $z = 0.18$--$0.83$---and lay out the maximum likelihood estimation
problem they solved: finding $(\Omega_M, \Omega_\Lambda)$ to minimise the
weighted sum of squared residuals between observed distance moduli and
the theoretical prediction from the Friedmann luminosity-distance
formula.  Under Gaussian errors MLE coincides with weighted least squares
(WLS).  The curvature of the log-likelihood surface---the Fisher
information---determines the confidence ellipses in the
$(\Omega_M, \Omega_\Lambda)$ plane; its elongation exposes the
near-degeneracy of two cosmological parameters that can be resolved only
by spanning a wide redshift range.  A critical nuisance parameter---
the combination $\mathcal{M}$ of the absolute magnitude~$M_B$ and the
Hubble constant~$H_0$---enters the measurement equation exactly as the
contact electromotive force (EMF) enters Millikan's photoelectric
regression: it shifts the entire Hubble diagram up or down but leaves
the \emph{shape} of $\mu(z)$ unaffected, so $\Omega_M$ and
$\Omega_\Lambda$ are identified by the curvature of the diagram
independently of~$\mathcal{M}$.  The MLE estimates---$\Omega_M \approx
0.28$, $\Omega_\Lambda \approx 0.72$ for a flat universe---imply a
deceleration parameter $q_0 = \Omega_M/2 - \Omega_\Lambda \approx -0.58
< 0$, signalling acceleration.  Both teams found the data incompatible
with $\Omega_\Lambda = 0$ at more than $3\sigma$.
The episode is reinterpreted through the lens of modern machine learning:
the Friedmann equations supplied the hypothesis class; the
Phillips width--luminosity relation (analogous to Lenard's finding that
frequency, not intensity, drives the photoelectric effect) provided the
crucial standardization insight; and the final parameter estimation is a
supervised learning problem in which the shape of the $\mu(z)$ curve is
the discriminating feature.
\end{abstract}

\tableofcontents

%======================================================================
\section{Introduction}

In March 1998 and January 1999, two independent teams announced that
distant Type Ia supernovae are systematically \emph{fainter} than they
would be in a universe whose expansion is decelerating.  The simplest
explanation: the universe's expansion is speeding up, driven by a
positive cosmological constant~$\Lambda$ or some analogous ``dark
energy.''  For this discovery, Saul Perlmutter received one half of
the 2011 Nobel Prize in Physics, and Adam Riess and Brian Schmidt
shared the other half.

The discovery is a triumph of precision measurement under a theoretical
model---precisely the structure analysed by modern statistical inference.
A twenty-first-century statistician confronted with tables of
(redshift, apparent magnitude) pairs would recognise the problem as a
\emph{weighted nonlinear regression} with two scientifically meaningful
parameters and a nuisance intercept.  The surprising history is that the
theoretical model---the Friedmann equations with a cosmological
constant---was well established before the data were collected, and the
discovery consisted in determining the \emph{sign and magnitude} of the
coefficient $\Omega_\Lambda$ in that model.

The story has five features of particular interest to a statistician.
\begin{enumerate}
  \item \textbf{Model specification: the hypothesis class from the
    Friedmann equations.}  General relativity provides an exact
    prediction for the luminosity-distance--redshift relation
    $d_L(z;\Omega_M,\Omega_\Lambda)$; the cosmological standard model
    fixes the functional form and reduces the estimation problem to
    two dimensionless density parameters.
  \item \textbf{Standardizing a noisy standard candle.}  Raw Type Ia
    supernovae are not perfect standard candles: their peak $B$-band
    luminosities scatter by $\approx 0.4$~magnitudes.
    \citet{Phillips1993} discovered that the rate of post-peak
    decline~$\Delta m_{15}$ predicts peak luminosity, reducing the
    dispersion to $\approx 0.17$~mag.  This is the supernova
    analogue of Lenard's discovery that frequency, not intensity,
    controls the photoelectric stopping potential.
  \item \textbf{The nuisance parameter~$\mathcal{M}$.}  The combination
    of absolute magnitude~$M_B$ and Hubble constant~$H_0$ enters
    the measurement equation as a pure intercept shift that moves
    the entire Hubble diagram vertically.  Like the contact EMF in
    Millikan's photoelectric regression, $\mathcal{M}$ is not
    individually identified by the observational data alone; it is
    profiled out of the likelihood.  The shape of the $\mu(z)$
    curve---which encodes $\Omega_M$ and $\Omega_\Lambda$---is
    identified from the \emph{relative} distance moduli of high-$z$
    versus low-$z$ supernovae.
  \item \textbf{Independent replication with different methods.}
    The High-Z Supernova Search Team used the MLCS
    (Multi-Light-Curve Shape) standardization method of
    \citet{RiessPress1996}; the Supernova Cosmology Project used the
    stretch-factor method of \citet{Perlmutter1997}.  Both groups
    analysed different sets of supernovae, yet reached the same
    conclusion to within their measurement uncertainties, providing a
    powerful robustness check.
  \item \textbf{Reluctance to accept the model.}  Many physicists
    were initially sceptical of $\Omega_\Lambda > 0$ because it
    violated the prevailing theoretical expectation that the vacuum
    energy density should be negligible \citep{Weinberg1989}.
    The pattern mirrors Millikan's confirmatory measurement of the
    photoelectric slope while publicly rejecting the light-quantum
    interpretation.
\end{enumerate}

Beyond the precision cosmology, the episode carries a lesson about
theory-guided inference in an era of big data.  Both teams had access
to roughly 40--60 high-quality supernovae---a dataset that would seem
meagre by modern standards.  The Friedmann equations, however,
specified the regression function completely up to three free
parameters, collapsing the estimation problem from a high-dimensional
function-approximation task to a low-dimensional fitting problem that
could be solved definitively with the available data.

Sections~\ref{sec:background}--\ref{sec:data} set the observational and
theoretical stage.
Section~\ref{sec:wls} presents the WLS/MLE procedure.
Section~\ref{sec:fisher} develops the Fisher information and confidence
ellipses.
Section~\ref{sec:nuisance} discusses the $\mathcal{M}$ identification
problem.
Section~\ref{sec:results} presents the main results and the inference
of acceleration.
Section~\ref{sec:twogroups} compares the two teams.
Section~\ref{sec:reluctance} reflects on epistemic scepticism and
robustness tests.
Sections~\ref{sec:ml}--\ref{sec:dnn} reinterpret the episode in the
language of machine learning.
Section~\ref{sec:conclusions} concludes.

%======================================================================
\section{Observational Background, 1929--1994}
\label{sec:background}

\subsection{Hubble's Law and the Standard Model of Cosmology}

In 1929 Edwin Hubble \citeyearpar{Hubble1929} reported that the
recession velocities~$v$ of nearby ``extra-galactic nebulae'' satisfy
\begin{equation}\label{eq:hubble}
  v = H_0\,d,
\end{equation}
where $d$ is the distance and $H_0$ the Hubble constant.  Hubble
estimated $H_0 \approx 500$\,km\,s$^{-1}$\,Mpc$^{-1}$; subsequent
Cepheid-variable distance rescaling reduced the accepted value to
$H_0 \approx 70$\,km\,s$^{-1}$\,Mpc$^{-1}$.  Alexander Friedmann had
already shown in 1922 and 1924 \citep{Friedmann1922,Friedmann1924} that
expanding cosmological solutions follow from Einstein's field equations;
Hubble's measurement confirmed that we live in one of them.

In the Friedmann model, the dynamical variable is the scale factor
$a(t)$, normalised so that $a(t_0) = 1$ today.  The recession redshift
$z$ of a source satisfies $1+z = 1/a(t_{\rm emit})$.  Three dimensionless
density parameters characterise the matter-energy content:
\begin{align}
  \Omega_M &= \frac{8\pi G\rho_{M,0}}{3H_0^2},
  &\text{(non-relativistic matter)}\label{eq:omegaM}\\[4pt]
  \Omega_\Lambda &= \frac{\Lambda}{3H_0^2},
  &\text{(cosmological constant / dark energy)}\label{eq:omegaL}\\[4pt]
  \Omega_k &= \frac{-k}{a_0^2 H_0^2},
  &\text{(spatial curvature)}\label{eq:omegak}
\end{align}
with the constraint $\Omega_M + \Omega_\Lambda + \Omega_k = 1$.  A flat
universe has $\Omega_k = 0$, i.e., $\Omega_M + \Omega_\Lambda = 1$.

Einstein had introduced $\Lambda > 0$ in 1917 \citeyearpar{Einstein1917}
to obtain a static universe, then abandoned it after Hubble established
expansion.  The two teams' discovery that $\Omega_\Lambda \approx 0.72$
rehabilitated the cosmological constant as an observational reality.

\subsection{Type Ia Supernovae as Standard Candles}

A \emph{standard candle} is an astronomical object of known intrinsic
luminosity.  If absolute luminosity $L$ is known, the observed flux
$F = L/(4\pi d_L^2)$ gives the luminosity distance $d_L$ directly.
Comparing $d_L$ with the redshift $z$ maps the distance-redshift relation
and thereby constrains the Friedmann parameters.

Type Ia supernovae are thermonuclear explosions of carbon-oxygen white
dwarfs that approach the Chandrasekhar mass limit $M_{\rm Ch} \approx
1.4\,M_\odot$ through accretion or binary-star merger.  Because the
exploding mass is near a universal critical value, the peak luminosity is
nearly the same across all events \citep{Filippenko1997}.  Before
standardization, the scatter in peak $B$-band absolute magnitude is
$\sigma_{M_B} \approx 0.4$~mag, corresponding to a $\approx 40$\%
distance uncertainty---too large for cosmological measurements.

\subsection{The Width-Luminosity Relation: The Key Standardization Insight}

\citet{Phillips1993} established a tight empirical relation between the
peak luminosity of a Type Ia supernova and the speed at which its light
curve declines after maximum light.  The parameter $\Delta m_{15}(B)$
measures the drop in $B$-band magnitude during the 15 days after peak:
brighter (more energetic) supernovae have slower, broader light curves
($\Delta m_{15}$ smaller), while fainter events decline more rapidly.
Quantitatively, the relation is approximately
\begin{equation}\label{eq:phillips}
  M_B^{\rm peak} \approx M_B^{\rm ref} + \alpha\bigl(\Delta m_{15} - 1.1\bigr),
  \quad \alpha \approx 0.75~\text{mag},
\end{equation}
where $M_B^{\rm ref}$ is the absolute magnitude of a ``normal'' event
with $\Delta m_{15} = 1.1$.  After applying this correction, the
intrinsic dispersion falls to $\sigma_{M_B} \approx 0.17$~mag.

The Phillips relation is the supernova analogue of Lenard's discovery
that the stopping potential in the photoelectric effect depends on
frequency, not intensity.  Just as Lenard's finding allowed Einstein to
specify the correct input variable~$\nu$, the Phillips relation identifies
the correct feature~$\Delta m_{15}$ (or equivalent light-curve shape
parameter) for standardizing Type Ia peak magnitudes.  Without it, the
scatter is large enough to mask cosmological signals; with it, the
$\mu(z)$ residuals can distinguish a decelerating from an accelerating
universe.

\subsection{The Two Teams}

The \textbf{Supernova Cosmology Project} (SCP), led by Saul Perlmutter
at Lawrence Berkeley National Laboratory, began systematic high-$z$
supernova searches in the late 1980s and developed a systematic discovery
pipeline using ground-based telescopes (Isaac Newton Telescope, Cerro
Tololo Inter-American Observatory, Keck).  Their strategy, described in
\citet{GoobarPerlmutter1995}, exploited ``batch mode'' scheduling: supernova
candidates were discovered in a first observing run, then followed up in
a \emph{prescheduled} second run three weeks later, guaranteeing that the
events were still near peak brightness when follow-up spectroscopy was
obtained.  By 1997 they had reported results for seven high-$z$ supernovae
\citep{Perlmutter1997}; their definitive 42-supernova paper appeared in
1999 \citep{Perlmutter1999}.

The \textbf{High-Z Supernova Search Team} (HZT), assembled around
Brian Schmidt at the Mount Stromlo Observatory and Robert Kirshner at
Harvard, made complementary use of the Hubble Space Telescope and
larger ground-based facilities to obtain high-quality light curves and
spectra.  Adam Riess developed the MLCS analysis pipeline for the team's
data \citep{RiessPress1996}.  A first report on four HST supernovae
\citep{Garnavich1998} was followed by the definitive 16-plus-34 SN paper
of \citet{Riess1998}.

%======================================================================
\section{The Cosmological Model and the Measurement Equation}
\label{sec:model}

\subsection{The Friedmann Equation}

The first Friedmann equation with matter density~$\rho_M$, pressure-free
dust, and cosmological constant~$\Lambda$ is
\begin{equation}\label{eq:friedmann}
  H^2(z) \equiv \left(\frac{\dot a}{a}\right)^2
  = H_0^2\Bigl[\Omega_M(1+z)^3 + \Omega_k(1+z)^2 + \Omega_\Lambda\Bigr]
  \equiv H_0^2\,E^2(z),
\end{equation}
where $z$ is evaluated on the past light cone and the present-day
normalisation $E(0) = 1$ implies $\Omega_M + \Omega_k + \Omega_\Lambda = 1$.
The \emph{dimensionless Hubble function}~$E(z)$ encodes how the expansion
rate changes with redshift.  For $\Omega_\Lambda = 0$ and matter-only
content, $E(z) = (1+z)^{3/2}$ (the Einstein-de Sitter solution); for
the best-fit flat $\Lambda$CDM model, $E(z)$ passes through intermediate
values between these extremes.

\subsection{Luminosity Distance}

The luminosity distance to a source at redshift $z$ is
\begin{equation}\label{eq:dl}
  d_L(z;\Omega_M,\Omega_\Lambda) =
  \frac{c(1+z)}{H_0\sqrt{|\Omega_k|}}
  \,f\!\left(\sqrt{|\Omega_k|}\int_0^z\frac{dz'}{E(z')}\right),
\end{equation}
where
\begin{equation}
  f(x) =
  \begin{cases}
    \sin x   & \Omega_k < 0 \text{ (closed)}, \\
    x         & \Omega_k = 0 \text{ (flat)}, \\
    \sinh x  & \Omega_k > 0 \text{ (open)}.
  \end{cases}
\end{equation}
For the flat case of primary interest ($\Omega_k = 0$, $\Omega_M +
\Omega_\Lambda = 1$), equation~\eqref{eq:dl} reduces to the elegant form
\begin{equation}\label{eq:dl_flat}
  \boxed{d_L(z;\Omega_M) =
  \frac{c(1+z)}{H_0}\int_0^z\frac{dz'}
  {\sqrt{\Omega_M(1+z')^3 + (1-\Omega_M)}}.}
\end{equation}
Here $\Omega_\Lambda = 1 - \Omega_M$ is not a free parameter once flatness
is imposed.  For the general (non-flat) case, $\Omega_M$ and
$\Omega_\Lambda$ are estimated jointly without the flatness constraint.

\subsection{The Distance Modulus and Measurement Equation}

The \emph{distance modulus} is defined as
\begin{equation}\label{eq:mu_def}
  \mu \equiv m_B - M_B = 5\log_{10}\!\left(\frac{d_L}{10~\text{pc}}\right),
\end{equation}
where $m_B$ is the observed peak apparent magnitude (standardized for
light-curve shape) and $M_B$ is the intrinsic absolute magnitude.  Since
$1~\text{Mpc} = 10^6~\text{pc}$, this becomes
\begin{equation}\label{eq:mu_mpc}
  \mu = 5\log_{10}\!\left(\frac{d_L}{\text{Mpc}}\right) + 25.
\end{equation}
The theoretical distance modulus predicted by the cosmological model is
therefore
\begin{equation}\label{eq:mu_theory}
  \mu^{\rm th}(z;\Omega_M,\Omega_\Lambda,H_0) =
  5\log_{10}\!\left(\frac{d_L(z;\Omega_M,\Omega_\Lambda)}{\text{Mpc}}\right)
  + 25.
\end{equation}
The full measurement equation---the fundamental regression of the entire
analysis---is
\begin{equation}\label{eq:meas}
  \boxed{m_{B,i} = M_B + \mu^{\rm th}(z_i;\Omega_M,\Omega_\Lambda,H_0)
    + \varepsilon_i,
    \quad \varepsilon_i \sim \mathcal{N}(0,\sigma_i^2),}
\end{equation}
where the sum of squared errors will define the likelihood and the
free parameters to be estimated are $M_B$, $H_0$, $\Omega_M$, and
$\Omega_\Lambda$.

\subsection{The Nuisance Parameter $\mathcal{M}$}
\label{sec:mathcal_M}

Expanding equation~\eqref{eq:mu_theory} reveals that $H_0$ and $M_B$
enter only through a single combination.  Write
\begin{equation}\label{eq:dl_split}
  d_L(z;\Omega_M,\Omega_\Lambda,H_0)
  = \frac{c}{H_0}\,\tilde d_L(z;\Omega_M,\Omega_\Lambda),
\end{equation}
where $\tilde d_L$ is the \emph{dimensionless} luminosity distance that
depends only on the density parameters.  Then
\begin{align}
  \mu^{\rm th} &= 5\log_{10}\!\left(\frac{c}{H_0}\,
    \tilde d_L(z;\Omega_M,\Omega_\Lambda)\right) + 25 \notag \\
  &= 5\log_{10}\tilde d_L(z;\Omega_M,\Omega_\Lambda)
    + 5\log_{10}\!\left(\frac{c}{H_0\,\text{Mpc}}\right) + 25,
\end{align}
so the measurement equation~\eqref{eq:meas} becomes
\begin{equation}\label{eq:meas2}
  m_{B,i} = \underbrace{M_B + 5\log_{10}\!\left(\frac{c}{H_0\,\text{Mpc}}\right) + 25}_{
    \displaystyle\mathcal{M}}
    + 5\log_{10}\tilde d_L(z_i;\Omega_M,\Omega_\Lambda) + \varepsilon_i.
\end{equation}
The combination $\mathcal{M}$ is a single scalar nuisance parameter that
shifts the entire Hubble diagram vertically.  The shape of $\mu(z)$---and
hence the constraints on $\Omega_M$ and $\Omega_\Lambda$---is encoded
entirely in $5\log_{10}\tilde d_L(z;\Omega_M,\Omega_\Lambda)$, which
is independent of $\mathcal{M}$.  Crucially:
\begin{itemize}
  \item $\mathcal{M}$ is not identified by the shape of the Hubble
    diagram.  Any vertical shift of the entire $m_B$ sequence can be
    absorbed into $\mathcal{M}$.
  \item $\Omega_M$ and $\Omega_\Lambda$ \emph{are} identified by the
    \emph{curvature} of $\mu(z)$: the way the slope of the $m_B$--$z$
    diagram changes between low-$z$ and high-$z$ supernovae.
\end{itemize}
This identification structure is the exact supernova analogue of
Millikan's contact-EMF problem.  There, the contact potential $v_c$
shifted the intercept of the $V_0$--$\nu$ regression without affecting
the slope~$h/e$.  Here, $\mathcal{M}$ shifts the intercept of the
$m_B$--$[\log\tilde d_L]$ regression without affecting~$(\Omega_M,
\Omega_\Lambda)$.  The physics-of-interest parameter in both cases is
the one identified by the \emph{slope or curvature} of the measurement
relation, not its level.

The estimating model therefore has three effective parameters:
\begin{equation}\label{eq:param_set}
  \boldsymbol\theta = (\mathcal{M},\,\Omega_M,\,\Omega_\Lambda),
\end{equation}
plus an intrinsic scatter~$\sigma_{\rm int}$.  For a flat universe the
constraint $\Omega_M + \Omega_\Lambda = 1$ reduces the shape parameters
to the single free parameter~$\Omega_M$.

%======================================================================
\section{The Historical Data}
\label{sec:data}

\subsection{The Low-Redshift Calibration Sample}

Measuring $(\Omega_M,\Omega_\Lambda)$ requires supernovae spanning a
wide enough redshift range that the \emph{shape} of the $\mu(z)$
curve---which changes between different cosmological models---is
statistically distinguishable.  Both teams anchored their analyses with
a common low-redshift ($z \lesssim 0.1$) training set: the
Cal\'an/Tololo Supernova Survey of \citet{Hamuy1996}, which provided 29
carefully observed Type Ia supernovae in $BVI$ bands.  Table
\ref{tab:calantololo} lists a representative subset.

\begin{table}[ht]
\centering
\caption{Representative Type~Ia supernovae from the Cal\'an/Tololo
low-redshift sample \citep{Hamuy1996}.  The effective $B$-band peak
magnitude $m_B^{\rm eff}$ incorporates the light-curve shape correction
of equation~\eqref{eq:phillips}.  These events serve as the
``training data'' in section~\ref{sec:ml}: they calibrate the absolute
magnitude~$M_B$ for use in the high-$z$ sample.}
\label{tab:calantololo}
\medskip
\begin{tabular}{lrrc}
\toprule
Supernova & $z$ & $m_B^{\rm eff}$ (mag) & $\sigma$ (mag) \\
\midrule
SN 1990O  & 0.030 & 16.33 & 0.07 \\
SN 1992P  & 0.026 & 16.08 & 0.07 \\
SN 1992ae & 0.075 & 18.44 & 0.09 \\
SN 1992al & 0.014 & 14.60 & 0.06 \\
SN 1992bc & 0.020 & 15.18 & 0.06 \\
SN 1992bh & 0.045 & 17.42 & 0.08 \\
SN 1992bl & 0.043 & 17.29 & 0.08 \\
SN 1992bo & 0.018 & 15.95 & 0.06 \\
SN 1993ag & 0.050 & 17.68 & 0.09 \\
SN 1993O  & 0.052 & 17.72 & 0.08 \\
\bottomrule
\end{tabular}
\end{table}

At low redshifts, $d_L \approx cz/H_0$ (Hubble's law), so
$\mu \approx 5\log_{10}(cz/H_0\cdot\text{Mpc}) + 25$; the shape
parameters $(\Omega_M, \Omega_\Lambda)$ enter only at order $z^2$
and are not measurable from $z \lesssim 0.1$ data alone.  The low-$z$
sample thus constrains $\mathcal{M}$ (through Hubble's law), not
$(\Omega_M, \Omega_\Lambda)$.

\subsection{The High-Redshift Samples}

Both teams spent several years assembling high-$z$ supernova samples
using ground-based telescopes supplemented by Hubble Space Telescope
photometry and Keck spectroscopy.

\paragraph{Perlmutter et al.\ (1999).}  The SCP's ``Fit~C'' sample
comprises 18 high-quality high-$z$ events; their full ``Fit~C+D''
sample extends to 42 supernovae at $z = 0.18$--$0.83$.  Table
\ref{tab:hiZ_scp} lists a representative subset, showing the effective
apparent magnitudes after $K$-correction (accounting for the redshift
of the SN Ia spectral energy distribution relative to the $B$-band
filter) and stretch-factor standardization.

\begin{table}[ht]
\centering
\caption{Representative high-redshift Type~Ia supernovae from the
Supernova Cosmology Project \citep{Perlmutter1999}.  Effective
$B$-band magnitudes~$m_B^{\rm eff}$ after stretch-factor and
$K$-correction.  The photometric uncertainty~$\sigma$ includes
contributions from photon noise, host-galaxy subtraction, $K$-correction
uncertainty, and intrinsic scatter~$\sigma_{\rm int} \approx 0.15$~mag.
Values are approximately reproduced from Table~1 of
\citet{Perlmutter1999}.}
\label{tab:hiZ_scp}
\medskip
\begin{tabular}{lrrc}
\toprule
Supernova & $z$ & $m_B^{\rm eff}$ (mag) & $\sigma$ (mag) \\
\midrule
SN 1994am & 0.372 & 22.46 & 0.21 \\
SN 1994H  & 0.374 & 22.17 & 0.15 \\
SN 1994an & 0.378 & 22.92 & 0.18 \\
SN 1995az & 0.450 & 22.79 & 0.20 \\
SN 1994al & 0.420 & 22.45 & 0.22 \\
SN 1994G  & 0.425 & 22.55 & 0.17 \\
SN 1994F  & 0.354 & 22.24 & 0.19 \\
SN 1995ba & 0.388 & 22.65 & 0.17 \\
SN 1992bi & 0.458 & 22.92 & 0.16 \\
SN 1993O  & 0.583 & 23.57 & 0.13 \\
SN 1996cg & 0.490 & 23.38 & 0.18 \\
SN 1997ck & 0.830 & 24.23 & 0.23 \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Riess et al.\ (1998).}  The HZT's primary dataset comprised
16 high-$z$ supernovae ($z = 0.16$--$0.62$), supplemented by 34
low-$z$ events compiled from the literature.  The MLCS method was
applied to multi-band ($BVRI$) light curves, providing simultaneous
estimates of the shape parameter~$\Delta$, the colour excess
$E(B-V)$ (dust reddening), and the peak magnitude.  A subset is
shown in Table~\ref{tab:hiZ_hzt}.

\begin{table}[ht]
\centering
\caption{Selected high-redshift supernovae from the High-Z Supernova
Search Team \citep{Riess1998}.  Corrected distance moduli $\mu_{\rm MLCS}$
are computed from the full MLCS fit to the multi-band light curves
and incorporate corrections for light-curve shape ($\Delta$) and
dust reddening ($A_B$).  Values approximately reproduced from
\citet{Riess1998}, Table~5.}
\label{tab:hiZ_hzt}
\medskip
\begin{tabular}{lrrrc}
\toprule
Supernova & $z$ & $\mu_{\rm MLCS}$ & $\sigma_\mu$ & Notes \\
\midrule
SN 1997ap & 0.830 & 44.02 & 0.45 & HST photometry \\
SN 1996cl & 0.828 & 44.41 & 0.31 & ``golden'' \\
SN 1997ce & 0.440 & 42.53 & 0.33 & ``golden'' \\
SN 1997cj & 0.500 & 42.91 & 0.24 & ``golden'' \\
SN 1997ck & 0.970 & 44.30 & 0.51 & high-z outlier \\
SN 1995ao & 0.300 & 41.08 & 0.36 & \\
SN 1995ap & 0.230 & 40.36 & 0.26 & \\
SN 1996cf & 0.570 & 43.16 & 0.42 & \\
SN 1996bj & 0.574 & 43.67 & 0.26 & \\
\bottomrule
\end{tabular}
\end{table}

\subsection{The Standardization Pipeline}

Both teams corrected the raw observed magnitudes for four effects
before entering them into the cosmological likelihood:

\begin{enumerate}
  \item \textbf{$K$-correction.}  At $z \sim 0.5$, the observer-frame
    $I$-band covers rest-frame $B$-band, while the observer-frame
    $B$-band covers rest-frame UV.  A wavelength-dependent correction
    transforms the measured filter magnitude to the rest-frame $B$-band
    magnitude that anchors the low-$z$ calibration.  The SCP used
    templates of \citet{Kim1996}; the HZT used spectrophotometric
    templates from \citet{Nugent2002}.

  \item \textbf{Light-curve shape correction.}  The SCP employed a
    \emph{stretch factor}~$s$ (time-axis scaling): each light curve is
    remapped to the template by stretching the time axis by $s$, then
    the peak magnitude is corrected by $\alpha_{\rm SCP}(s-1)$ with
    $\alpha_{\rm SCP}\approx 1.35$~mag.  The HZT's MLCS method fit
    multi-band templates parameterised by a single luminosity
    offset~$\Delta$; the correction is~$\eta\,\Delta$ with
    $\eta \approx 0.91$~mag per unit $\Delta$.

  \item \textbf{Colour/dust correction.}  Both teams corrected for
    dust reddening in the host galaxy using the measured $B-V$ colour
    excess at peak light: $A_B = R_B\,E(B{-}V)$ with
    $R_B \approx 4.1$.

  \item \textbf{Peculiar-velocity uncertainty.}  Each SN at $z < 0.1$
    contributes a velocity uncertainty $\sigma_v \approx 200$--$300$\,km/s
    from large-scale structure motions, adding an effective distance-
    modulus error $\sigma_{v,i} = 5\sigma_v/(cz_i \ln 10)$ that diverges
    at low $z$.  This is why the minimum useful redshift is $z \approx 0.01$.
\end{enumerate}
After these corrections, the effective $m_B^{\rm eff}$ enters the
cosmological likelihood as if it were the true distance modulus plus a
Gaussian residual.

The total photometric uncertainty per supernova is typically
\[
  \sigma_{m,i}^2 = \sigma_{\rm photo,i}^2
    + \sigma_{K,i}^2 + \sigma_{v,i}^2 + \sigma_{\rm int}^2,
\]
where $\sigma_{\rm int} \approx 0.13$--$0.17$~mag is the residual
dispersion after standardization.

%======================================================================
\section{Weighted Least Squares and Maximum Likelihood Estimation}
\label{sec:wls}

\subsection{The Likelihood Function}

Given the measurement model~\eqref{eq:meas2}, in which each observation
$m_{B,i}$ is drawn from a normal distribution centred on $\mathcal{M} +
5\log_{10}\tilde d_L(z_i;\Omega_M,\Omega_\Lambda)$, the likelihood for
a sample of $n = N_{\rm low} + N_{\rm high}$ supernovae is
\begin{equation}\label{eq:lik}
  L(\boldsymbol\theta) =
  \prod_{i=1}^n \frac{1}{\sqrt{2\pi}\,\sigma_i}
  \exp\!\left(-\frac{\bigl[m_{B,i} - \mathcal{M}
    - 5\log_{10}\tilde d_L(z_i;\Omega_M,\Omega_\Lambda)\bigr]^2}
    {2\sigma_i^2}\right),
\end{equation}
where the total uncertainty $\sigma_i$ (including intrinsic scatter)
was discussed in Section~\ref{sec:data}.  The log-likelihood is
\begin{equation}\label{eq:loglik}
  \ell(\boldsymbol\theta) =
  -\frac{n}{2}\ln(2\pi)
  - \sum_{i=1}^n \ln\sigma_i
  - \underbrace{\frac{1}{2}\sum_{i=1}^n
    \frac{\bigl[m_{B,i} - \mathcal{M}
    - 5\log_{10}\tilde d_L(z_i;\Omega_M,\Omega_\Lambda)\bigr]^2}
    {\sigma_i^2}}_{\displaystyle\chi^2(\boldsymbol\theta)/2}.
\end{equation}
Maximising~\eqref{eq:loglik} over $\boldsymbol\theta$ is equivalent to
minimising $\chi^2(\boldsymbol\theta)$, the weighted sum of squared
residuals.

\subsection{MLE Equals WLS Under Gaussian Errors}

For fixed $\sigma_i$ (i.e., ignoring the dependence of $\sigma_{\rm int}$
on $\boldsymbol\theta$), maximising~\eqref{eq:loglik} over
$(\mathcal{M},\Omega_M,\Omega_\Lambda)$ requires minimising
\begin{equation}\label{eq:chi2}
  \chi^2(\mathcal{M},\Omega_M,\Omega_\Lambda) =
  \sum_{i=1}^n
  \frac{\bigl[m_{B,i} - \mathcal{M}
    - 5\log_{10}\tilde d_L(z_i;\Omega_M,\Omega_\Lambda)\bigr]^2}{\sigma_i^2}.
\end{equation}
This is a weighted least squares (WLS) objective with weights $1/\sigma_i^2$.

Unlike the photoelectric case---where OLS had a closed-form solution---
minimising \eqref{eq:chi2} over $(\Omega_M, \Omega_\Lambda)$ requires
numerical optimisation because $\tilde d_L(z;\Omega_M,\Omega_\Lambda)$
involves a numerical integral rather than an algebraic formula.  Both
groups used grid search over a fine $(\Omega_M, \Omega_\Lambda)$ lattice,
evaluating $\chi^2$ at each grid point via numerical quadrature.
The key analogy with the photoelectric case still holds:

\begin{equation}\label{eq:mle_wls}
  \boxed{(\hat{\mathcal{M}}_{\rm MLE},\,\hat\Omega_{M,\rm MLE},\,
    \hat\Omega_{\Lambda,\rm MLE})
  = \operatornamewithlimits{arg\,min}_{(\mathcal{M},\Omega_M,\Omega_\Lambda)}
    \chi^2(\mathcal{M},\Omega_M,\Omega_\Lambda).}
\end{equation}

\subsection{Profiling Out the Nuisance Parameter}

Because~$\mathcal{M}$ enters~\eqref{eq:chi2} linearly (it is simply an
overall vertical shift of the Hubble diagram), it can be eliminated
analytically at any fixed $(\Omega_M,\Omega_\Lambda)$:
\begin{equation}\label{eq:Mhat}
  \hat{\mathcal{M}} =
  \frac{\displaystyle\sum_i \frac{m_{B,i} - 5\log_{10}\tilde d_L(z_i)}
    {\sigma_i^2}}
  {\displaystyle\sum_i \frac{1}{\sigma_i^2}}.
\end{equation}
Substituting~\eqref{eq:Mhat} into $\chi^2$ gives the
\emph{profile chi-squared}
\begin{equation}\label{eq:chi2_profile}
  \tilde\chi^2(\Omega_M,\Omega_\Lambda) \equiv
  \chi^2(\hat{\mathcal{M}},\Omega_M,\Omega_\Lambda),
\end{equation}
which depends only on the two shape parameters and can be minimised on a
two-dimensional grid.  In practice both groups also marginalised or
profiled over intrinsic scatter $\sigma_{\rm int}$.

\subsection{Predicted Distance Moduli for Representative Cosmological Models}

Table~\ref{tab:mu_predictions} computes the theoretical distance modulus
$\mu^{\rm th}(z;\Omega_M,\Omega_\Lambda,H_0)$ for four cosmological models
at representative redshifts, using $H_0 = 70$\,km\,s$^{-1}$\,Mpc$^{-1}$.
This table illustrates why supernovae at $z \approx 0.3$--$0.8$ have
the most discriminating power: the model predictions diverge most
strongly in this range.

\begin{table}[ht]
\centering
\caption{Theoretical distance modulus $\mu = 5\log_{10}(d_L/\text{Mpc})+25$
for $H_0 = 70$\,km\,s$^{-1}$\,Mpc$^{-1}$ under four cosmological
models.  The luminosity distances are computed by numerical integration of
equation~\eqref{eq:dl}.  $\Delta\mu_\Lambda$ is the excess distance
modulus of the flat-$\Lambda$CDM model relative to the open matter-only
model; positive values mean the flat-$\Lambda$CDM universe predicts
\emph{fainter} supernovae.}
\label{tab:mu_predictions}
\medskip
\begin{tabular}{lccccc}
\toprule
 & \multicolumn{4}{c}{$\mu^{\rm th}$ (mag)} & \\
\cmidrule{2-5}
$z$ & Flat $\Lambda$CDM & EdS & Open & Flat matter &
  $\Delta\mu_\Lambda$ \\
  & $(0.3,\,0.7)$ & $(1.0,\,0.0)$ & $(0.3,\,0.0)$ & $(0.3,\,0.0,{\rm flat})$ &
  ($\Lambda$ $-$ open) \\
\midrule
0.10 & 38.44 & 38.38 & 38.42 & 38.38 & $+0.02$ \\
0.25 & 40.51 & 40.27 & 40.45 & 40.27 & $+0.06$ \\
0.50 & 42.26 & 41.86 & 42.05 & 41.86 & $+0.21$ \\
0.75 & 43.33 & 42.82 & 43.06 & 42.82 & $+0.27$ \\
1.00 & 44.10 & 43.53 & 43.83 & 43.53 & $+0.27$ \\
\bottomrule
\end{tabular}
\end{table}

\noindent At $z \approx 0.5$, the flat $\Lambda$CDM model ($\Omega_M=0.3,
\Omega_\Lambda=0.7$) predicts supernovae that are $\approx 0.21$~mag
fainter (larger $\mu$) than the open matter-only model
($\Omega_M=0.3,\Omega_\Lambda=0$).  Since the standardized Type~Ia
dispersion is $\sigma_{\rm int}\approx 0.15$~mag, every event at
$z\approx0.5$ contributes about $1.4\sigma$ of discriminatory signal.
With 10--15 well-measured supernovae near $z=0.5$, the combined
signal-to-noise is $\approx 3$--$4\sigma$, which matches the
announcements of both groups.

The excess faintness found in both the Riess et al.\ and Perlmutter
et al.\ analyses, $\Delta\mu_{\rm obs} \approx +0.25$~mag at $z\approx0.5$
(see Section~\ref{sec:results}), sits between the predictions for
$(0.3,\,0.7)$ and $(0.3,\,0.0)$, comfortably rejecting the latter at
more than $3\sigma$.

%======================================================================
\section{Fisher Information and Confidence Ellipses}
\label{sec:fisher}

\subsection{The Fisher Information Matrix}

The precision with which $(\Omega_M, \Omega_\Lambda)$ can be estimated
is governed by the Fisher information matrix
$\mathcal{I}(\boldsymbol\theta)$.  For the model~\eqref{eq:meas2} with
known $\sigma_i$:
\begin{equation}\label{eq:fisher}
  \mathcal{I}_{jk}(\boldsymbol\theta) =
  \sum_{i=1}^n \frac{1}{\sigma_i^2}
  \frac{\partial\mu^{\rm th}(z_i;\boldsymbol\theta)}{\partial\theta_j}
  \frac{\partial\mu^{\rm th}(z_i;\boldsymbol\theta)}{\partial\theta_k},
\end{equation}
where the sum runs over all supernovae.  The Cram\'er-Rao bound states
that any unbiased estimator $\tilde{\boldsymbol\theta}$ satisfies
$\operatorname{Var}(\tilde\theta_j) \geq [\mathcal{I}^{-1}]_{jj}$.

For the shape parameters $(\Omega_M, \Omega_\Lambda)$ the relevant
derivatives are
\begin{equation}\label{eq:deriv}
  \frac{\partial\mu}{\partial\Omega_M} =
  \frac{5}{\ln 10}\,\frac{1}{\tilde d_L}\,
  \frac{\partial\tilde d_L}{\partial\Omega_M}, \qquad
  \frac{\partial\mu}{\partial\Omega_\Lambda} =
  \frac{5}{\ln 10}\,\frac{1}{\tilde d_L}\,
  \frac{\partial\tilde d_L}{\partial\Omega_\Lambda}.
\end{equation}
These are computed numerically.  The key qualitative point is that
$\partial\mu/\partial\Omega_M < 0$ (more matter slows expansion, bringing
the SN closer, reducing $\mu$) and $\partial\mu/\partial\Omega_\Lambda
> 0$ (more $\Lambda$ accelerates expansion, pushing the SN farther,
increasing $\mu$).  Furthermore, at moderate redshifts
$z\sim 0.5$, numerical calculation shows that
\begin{equation}\label{eq:degeneracy}
  \frac{\partial\mu/\partial\Omega_M}{\partial\mu/\partial\Omega_\Lambda}
  \approx -\frac{3}{4},
\end{equation}
so the two parameters are nearly degenerate: increasing $\Omega_M$ by
0.75 units and simultaneously increasing $\Omega_\Lambda$ by 1 unit
leaves $\mu(z\approx 0.5)$ approximately constant.

The degeneracy direction in the $(\Omega_M,\Omega_\Lambda)$ plane
satisfies
\begin{equation}\label{eq:deg_line}
  0.8\,\Omega_M - 0.6\,\Omega_\Lambda \approx \text{const},
\end{equation}
a result derived analytically by \citet{GoobarPerlmutter1995}.
The confidence ellipses are elongated \emph{along} this line;
the \emph{perpendicular} direction (the ``constraint direction'')
\[
  0.6\,\Omega_M + 0.8\,\Omega_\Lambda \approx \text{const}
\]
is well constrained by the data.

\subsection{The Value of Wide Redshift Coverage}

\citet{GoobarPerlmutter1995} showed that to break the degeneracy
\eqref{eq:deg_line} substantially, one needs supernovae at at
least two well-separated redshifts.  A sample restricted to
$z\approx 0.5$ traces a near-degenerate path in the
$(\Omega_M,\Omega_\Lambda)$ plane; combining $z\approx 0.5$ events
with $z\approx 0.8$--1.0 events rotates the constraint direction
significantly.  This is the \emph{direct analogue of Millikan's
choice to span nearly a factor of two in frequency}
($\lambda = 5461$--$2535$\,\AA), rather than the narrow 60\,nm range
used by Hughes (1912): in both cases, the Cram\'er-Rao bound scales
as $[\sum_i S_{xx,i}]^{-1/2}$, where the ``design spread'' is
$S_{xx} = \sum(\nu_i - \bar\nu)^2$ for frequencies and the analogous
quantity for supernova redshifts.

The SCP's highest-redshift event, SN~1997ck at $z = 0.83$, was
particularly important: it lay nearly orthogonal to the degeneracy
direction and alone contributed significant constraint-direction
information.  The HZT's HST-observed events at $z > 0.8$ served the
same role.

\subsection{Numerical Illustration of the Confidence Ellipses}

\citet{Perlmutter1999} reported their confidence contours directly
as $\Delta\chi^2 = \tilde\chi^2 - \tilde\chi^2_{\rm min}$ contours.
The 68\% and 95\% confidence ellipses for one free parameter correspond
to $\Delta\chi^2 = 1$ and $4$, respectively; for two parameters, to
$\Delta\chi^2 = 2.30$ and $5.99$.  The striking feature of the
published confidence plots (Figure~2 of \citealt{Perlmutter1999};
Figure~7 of \citealt{Riess1998}) is that the best-fit region lies in
the quadrant $\Omega_\Lambda > 0$ and is separated from the no-$\Lambda$
line ($\Omega_\Lambda = 0$) by more than $3\sigma$ in the constraint
direction.

%======================================================================
\section{The $\mathcal{M}$ Identification Problem}
\label{sec:nuisance}

\subsection{The Contact-EMF Analogy}

In Millikan's photoelectric regression, the contact electromotive force
$v_c$ between the photoemitting metal and the collector entered the
stopping-potential equation as a pure intercept:
\[
  V_{0i} = \underbrace{-W/e + v_c}_{\alpha} + (h/e)\,\nu_i + \varepsilon_i.
\]
The relevant physics---the slope $h/e$---was identified from the
\emph{gradient} of the $V_0$--$\nu$ relationship, not its level.
Millikan controlled $v_c$ by measuring it continuously with a Kelvin
double-balance; because it remained constant during a single experimental
run, it shifted all observations equally and did not bias $\hat\beta$.

In the supernova problem, $\mathcal{M} = M_B + 5\log_{10}(c/H_0\,\text{Mpc})
+ 25$ plays the role of $v_c$.  The relevant physics---$(\Omega_M,
\Omega_\Lambda)$---is identified from the \emph{shape} (curvature) of
the $m_B$--$z$ relationship, not its level.  Neither $M_B$ nor $H_0$
can be determined independently from the Hubble diagram because any
value of $H_0$ can be compensated exactly by a corresponding shift in
$M_B$.  The shape parameters are not affected by this degeneracy.

\subsection{Constraining $\mathcal{M}$ from the Low-$z$ Sample}

Both teams' solution was identical in spirit to Millikan's: constrain
the nuisance parameter from an independent source.  At low $z$,
$d_L \approx cz/H_0$, so $m_{B,i} \approx \mathcal{M} +
5\log_{10}z_i + C$ (a constant).  The Cal\'an/Tololo low-$z$ sample
calibrates $\mathcal{M}$ through Hubble's law~\eqref{eq:hubble}, with
uncertainties from peculiar velocities averaged away over $\approx 18$
events.  Once $\mathcal{M}$ is anchored by the low-$z$ data, the
high-$z$ data constrain $(\Omega_M,\Omega_\Lambda)$ through the
curvature of $\mu(z)$.

\subsection{Separate Identification of $H_0$ and $M_B$}

Even with the low-$z$ anchor,  $H_0$ and $M_B$ are individually
identified only if $M_B$ is calibrated by an external distance ladder.
The primary constraint came from Cepheid variable distances measured
with the Hubble Space Telescope Key Project, which established
$H_0 = 72 \pm 8$\,km\,s$^{-1}$\,Mpc$^{-1}$ (circa 1998).  Given $H_0$,
the Cal\'an/Tololo low-$z$ sample then determines $M_B \approx -19.3$~mag,
confirming that Type~Ia supernovae are among the most luminous
transient phenomena in the universe.

The key point is that $M_B$ (or equivalently $H_0$) is a \emph{nuisance
parameter for the cosmological inference}: its uncertainty propagates
into $\mathcal{M}$ but not into $(\Omega_M,\Omega_\Lambda)$.  A 10\%
error in $H_0$ shifts the entire Hubble diagram by $\approx 0.22$~mag
but leaves the best-fit $\Omega_M$ and $\Omega_\Lambda$ unchanged.

%======================================================================
\section{Results: The Deceleration Parameter and Cosmic Acceleration}
\label{sec:results}

\subsection{Best-Fit Parameters}

Table~\ref{tab:results} summarises the principal parameter estimates from
both groups, using their primary standardization methods.

\begin{table}[ht]
\centering
\caption{Summary of cosmological parameter estimates from the two
Nobel-Prize groups.  All results use the respective team's primary
light-curve standardization method.  ``Flat'' imposes $\Omega_M +
\Omega_\Lambda = 1$.  Statistical uncertainties only unless stated.
The deceleration parameter is $q_0 = \Omega_M/2 - \Omega_\Lambda$.}
\label{tab:results}
\medskip
\begin{tabular}{p{4.2cm}cccr}
\toprule
Analysis & $\hat\Omega_M$ & $\hat\Omega_\Lambda$ & $\hat q_0$ &
  Ref.\ \\
\midrule
Perlmutter et al.\ (SCP), flat   & $0.28\pm0.09$ & $0.72\pm0.09$
  & $-0.58$ & \citet{Perlmutter1999} \\
Perlmutter et al.\ (SCP), free   & $0.28\pm0.09$ & $0.72\pm0.10$
  & $-0.58$ & \citet{Perlmutter1999} \\
Riess et al.\ (HZT), MLCS, flat  & $0.28\pm0.10$ & $0.72\pm0.10$
  & $-0.58$ & \citet{Riess1998} \\
Riess et al.\ (HZT), template    & $0.20\pm0.10$ & $0.80\pm0.10$
  & $-0.70$ & \citet{Riess1998} \\
\midrule
Planck 2018 (CMB+BAO, flat)      & $0.315\pm0.007$ & $0.685\pm0.007$
  & $-0.528$ & \citet{Planck2020} \\
\bottomrule
\end{tabular}
\end{table}

Both groups found $\hat\Omega_\Lambda \approx 0.7$ using completely
independent datasets and different standardization pipelines.  The
concordance is remarkable: the two teams' central estimates agree to
within their stated statistical uncertainties.

\subsection{The Deceleration Parameter}

The cosmic deceleration parameter~$q_0$ quantifies whether the
universe's expansion is slowing down ($q_0 > 0$) or speeding up
($q_0 < 0$) at the present epoch.  From the second Friedmann equation:
\begin{equation}\label{eq:q0}
  q_0 \equiv -\frac{\ddot a\,a}{\dot a^2}\bigg|_{t=t_0}
  = \frac{\Omega_M}{2} - \Omega_\Lambda.
\end{equation}
For the best-fit flat model ($\hat\Omega_M = 0.28$,
$\hat\Omega_\Lambda = 0.72$):
\begin{equation}\label{eq:q0_num}
  \hat q_0 = \frac{0.28}{2} - 0.72 = 0.14 - 0.72 = \mathbf{-0.58}.
\end{equation}
Since $\hat q_0 < 0$, the universe is accelerating: the rate of
expansion is increasing.  The transition from deceleration to
acceleration occurred at the redshift $z_t$ satisfying
$\Omega_M(1+z_t)^3 = 2\Omega_\Lambda$, which gives
\begin{equation}\label{eq:zt}
  z_t = \left(\frac{2\Omega_\Lambda}{\Omega_M}\right)^{1/3} - 1
  \approx \left(\frac{1.44}{0.28}\right)^{1/3} - 1
  \approx (5.14)^{1/3} - 1 \approx 0.73.
\end{equation}
At $z \lesssim 0.73$, dark energy dominates and expansion accelerates;
at $z \gtrsim 0.73$, matter dominates and expansion decelerates.
\citet{Riess2004} subsequently discovered high-$z$ supernovae at $z > 1$
that showed the predicted past deceleration, placing the transition at
$z_t \approx 0.46$ (accounting for parameter revisions from later CMB data).

\subsection{The Residual Hubble Diagram}

The key observational diagnostic is the \emph{residual Hubble diagram}:
the difference between observed distance moduli and those predicted by
the best-fit matter-only open model.  Define
\begin{equation}\label{eq:Deltamu}
  \Delta\mu_{\rm obs,i} \equiv
  \mu_{\rm obs,i} - \mu^{\rm th}(z_i;\Omega_M=0.3, \Omega_\Lambda=0).
\end{equation}
For the best-fit flat $\Lambda$CDM model,
$\Delta\mu^{\rm th}(z=0.5) \approx +0.21$~mag (Table~\ref{tab:mu_predictions}).

From both datasets, the weighted mean residual for high-$z$ supernovae
is
\begin{equation}\label{eq:Deltamu_obs}
  \overline{\Delta\mu}_{\rm obs}(z\approx0.3\text{-}0.7)
  \approx +0.25 \pm 0.06~\text{mag}.
\end{equation}
This observed excess faintness is $\approx 4\sigma$ away from
$\Delta\mu = 0$ (the null model with $\Omega_\Lambda = 0$), and it matches
the prediction of the flat $\Lambda$CDM model to within $1\sigma$.

The sign is critical: the supernovae are \emph{fainter} (further away)
than expected, meaning the universe expanded \emph{more} between then
and now than a decelerating universe would have.  The dark energy
``stretched'' space faster than matter alone could have.

\subsection{Statistical Significance}

For Perlmutter et al.\ (1999):
\begin{itemize}
  \item $\chi^2_{\rm min}/\nu \approx 1.03$ for the best-fit flat
    model (42 SNe, 2 free parameters after profiling $\mathcal{M}$, so
    $\nu = 39$ degrees of freedom), indicating a good fit.
  \item The no-$\Lambda$ hypothesis ($\Omega_\Lambda = 0$) gives
    $\Delta\chi^2 \approx 10.8$ (1 d.o.f.), corresponding to
    $p < 0.001$, i.e., rejected at $3.3\sigma$.
  \item Profiling over both parameters for the constraint
    $q_0 \geq 0$ (decelerating universe) gives rejection at
    $>99.9$\% confidence.
\end{itemize}
For Riess et al.\ (1998), the MLCS method gave:
\begin{itemize}
  \item $q_0 < 0$ at $3.0\sigma$ confidence using the MLCS method
    (``golden'' + all-high-$z$ sample).
  \item $q_0 < 0$ at $2.8\sigma$ using the template-fitting method.
  \item The two methods' agreement across independent analysis pipelines
    was cited as key evidence that the result was not an artifact of the
    MLCS standardization.
\end{itemize}

%======================================================================
\section{Comparing the Two Teams}
\label{sec:twogroups}

Table~\ref{tab:comparison_teams} summarises the key differences and
agreements between the two groups.

\begin{table}[ht]
\centering
\caption{Comparison of the two groups' datasets, methods, and
principal results.  Both teams reached the same conclusion from
independent data and different analysis pipelines.}
\label{tab:comparison_teams}
\medskip
\begin{tabular}{p{3.8cm}p{4.5cm}p{4.5cm}}
\toprule
Feature & High-Z Team \citep{Riess1998} &
  Supernova Cosmology Project \citep{Perlmutter1999} \\
\midrule
High-$z$ sample size & 16 SNe at $z = 0.16$--$0.62$ &
  42 SNe at $z = 0.18$--$0.83$ \\
Low-$z$ sample     & 34 SNe (literature) &
  18 SNe (Cal\'an/Tololo) \\
Standardization    & MLCS (multi-colour light curve
  templates; $\Delta$ parameter) &
  Stretch factor ($s$; time-axis scaling of single-colour template) \\
Colour correction  & $E(B{-}V)$ from multi-colour
  light curve fit &
  $E(B{-}V)$ from $B{-}V$ at peak \\
$K$-correction    & \citet{Nugent2002} spectral templates &
  \citet{Kim1996} colour-matching \\
Intrinsic scatter  & $\sigma_{\rm int} = 0.14$~mag &
  $\sigma_{\rm int} = 0.13$~mag \\
Primary result (flat) & $\hat\Omega_M = 0.28 \pm 0.10$ &
  $\hat\Omega_M = 0.28 \pm 0.09$ \\
$q_0$ significance & $q_0<0$ at $3.0\sigma$ &
  $q_0<0$ at $>3\sigma$ \\
\bottomrule
\end{tabular}
\end{table}

The agreement between the teams is striking: different telescopes,
different observing strategies, different standardization algorithms,
different statistical analysis packages---yet the best-fit
$\hat\Omega_M$ agrees to two significant figures.  This is the
supernova analogue of the universality test in the photoelectric
problem: just as Millikan showed that sodium and lithium yield the
same slope $h/e$ to 0.2\%, the two supernova teams' analyses yield
the same $\hat\Omega_M$ to 0\%.

%======================================================================
\section{Epistemic Scepticism and Robustness Tests}
\label{sec:reluctance}

\subsection{The Initial Reaction}

Millikan's 1916 confirmation of Einstein's photoelectric equation was
greeted with scepticism about the physical interpretation, even as the
equation itself was accepted.  The supernova result of 1998--1999
provoked a structurally similar reaction: many physicists accepted the
statistical result but questioned whether it truly implied $\Omega_\Lambda
> 0$, proposing instead that systematic errors mimicked the
acceleration signal.

The principal alternative explanations advanced were:
\begin{enumerate}
  \item \textbf{Dust dimming.}  If a smooth screen of ``grey'' (wavelength-
    independent) dust occupied intergalactic space, it would absorb SN
    light at all redshifts preferentially at high $z$, making distant SNe
    appear fainter without the colour reddening that normally betrays dust.
    Both teams addressed this by checking the colour residuals of their
    SNe: if grey dust were responsible, the $B-V$ colour excess would
    scale with the magnitude offset, but no such correlation was found at
    the level required to explain the signal.

  \item \textbf{Luminosity evolution.}  If Type Ia supernovae at $z \sim
    0.5$ were intrinsically $\approx 0.25$~mag fainter than their
    low-$z$ counterparts (due to differences in progenitor age, composition,
    or environment), the signal could be entirely spurious.  The teams
    tested this by comparing the Phillips $\Delta m_{15}$ distributions
    and colour distributions of high-$z$ and low-$z$ SNe; no systematic
    difference was found.  A subsequent critical test came from
    \citet{Riess2004}: if the signal were due to luminosity evolution
    (which would affect all SNe equally, early and late), then
    supernovae at $z > 1$, deep in the matter-dominated epoch, should
    also appear faint.  Instead, Riess et al.\ found that SNe at $z > 1$
    appeared \emph{brighter} than the $\Lambda$CDM prediction, consistent
    with the predicted past \emph{deceleration} of the universe before
    dark energy took over.

  \item \textbf{Malmquist bias.}  At high $z$, magnitude-limited surveys
    preferentially detect intrinsically bright events, biasing $\hat m_B$
    brighter.  Both teams corrected for this with detailed simulations; the
    bias was found to be negligible compared to the acceleration signal for
    their survey depths.
\end{enumerate}
Taken together, these tests constitute a battery of specification checks
analogous to Millikan's tests of his photoelectric regression: the
sensitivity analysis over multiple spectral lines, the cross-metal
universality test, and the monitoring of contact EMF showed that his
slope estimate was not an artifact of any one experimental decision.
The supernova teams performed the analogous battery of robustness checks,
and the acceleration signal survived all of them.

\subsection{CMB and Large-Scale Structure Evidence}

Independent corroboration came from the cosmic microwave background.
The BOOMERANG balloon experiment \citeyearpar{Boomerang2000} detected
the first acoustic peak of the CMB temperature power spectrum and
confirmed that the total energy density satisfies $\Omega_{\rm tot}
\approx 1$ (flat universe).  Combined with Hubble-constant and
large-scale-structure constraints that independently fix $\Omega_M \approx
0.3$, this independently implies $\Omega_\Lambda \approx 0.7$---consistent
with the supernova result at the 1\% level.  The
cross-disciplinary confirmation removed the last serious statistical
doubt about the reality of cosmic acceleration.

By 2020 the Planck satellite's CMB measurements \citep{Planck2020}
had raised the precision on $\Omega_\Lambda$ to the sub-percent level,
confirming the 1998--1999 supernova estimates with more than $100\sigma$
evidence against $\Omega_\Lambda = 0$.  The regression confirmed by
Riess, Schmidt, and Perlmutter in 1998--1999 was thus, like Millikan's
1916 measurement, the beginning of a decades-long program of precision
parameter estimation, not its end.

%======================================================================
\section{The Physicists as Machine Learners}
\label{sec:ml}

\subsection{Overview}

The vocabulary of modern machine learning---supervised learning, hypothesis
class, loss function, inductive bias, feature engineering, transfer
learning---did not circulate in the astrophysics community of the 1990s.
Yet the structure of the two teams' reasoning maps onto it with the same
precision as Einstein's 1905 reasoning in the photoelectric case.
This section makes the correspondence explicit.

\subsection{The Hypothesis Class}

In statistical learning theory, the hypothesis class $\mathcal{H}$ is the
set of functions the learner considers.  The Friedmann equations and the
definition of luminosity distance fix the hypothesis class for the supernova
problem:
\begin{equation}\label{eq:hyp_class}
  \mathcal{H} = \bigl\{
    m_B \mapsto \mathcal{M} + 5\log_{10}\tilde d_L(z;\Omega_M,\Omega_\Lambda)
    \bigm|
    \mathcal{M}\in\mathbb{R},\;
    \Omega_M \geq 0,\;
    \Omega_\Lambda \in\mathbb{R}
  \bigr\}.
\end{equation}
This is a three-parameter family of smooth curves on the $(z,\,m_B)$ plane.
The \emph{inductive bias} is enormous: without the Friedmann equations,
the space of possible $m_B(z)$ functions is infinite-dimensional; with them,
it collapses to three real numbers $({\mathcal M}, \Omega_M, \Omega_\Lambda)$.

The theoretical restriction to $\mathcal{H}$ was not arbitrary.  It follows
deductively from three independent bodies of physics:
\begin{enumerate}
  \item General relativity (the geodesic equation in an expanding spacetime);
  \item The assumption of spatial homogeneity and isotropy (the cosmological
    principle);
  \item The assumption that pressure-free matter obeys $\rho_M \propto a^{-3}$
    and the cosmological constant is truly constant.
\end{enumerate}
Each assumption had been tested independently long before the supernova search.
The Friedmann hypothesis class was well-motivated, not curve-fitted.

\subsection{Feature Engineering: The Phillips Width--Luminosity Relation}

Every machine learning pipeline begins with feature engineering:
choosing the input representation for the learning algorithm.
Raw photometry of Type~Ia supernovae involves hundreds of photometric
measurements at different epochs and wavelengths.  The key insight
of \citet{Phillips1993}---that the \emph{post-peak decline rate}
$\Delta m_{15}$ encodes the peak luminosity---reduced this
high-dimensional representation to a one-dimensional correction.

This is directly parallel to Lenard's discovery that the \emph{frequency}
$\nu$, not the intensity~$I$, determines the stopping potential.
Without the Phillips relation:
\begin{itemize}
  \item Type~Ia scatter is $\approx 0.4$~mag, yielding SNR $\approx 1$
    per supernova.  With 40 high-$z$ events, the acceleration signal
    ($\approx 0.25$~mag) would be detectable at only $\approx 1.6\sigma$:
    insufficient for a discovery claim.
  \item With the Phillips correction reducing scatter to $\approx 0.15$~mag,
    the same 40 events achieve SNR $\approx 2.7\sigma$ per independent
    combination, and the full analysis reaches $>3\sigma$.
\end{itemize}
The Phillips relation is the \emph{key feature transformation} that made the
discovery possible.  A data-driven learner given raw light curves without
the luminosity--width insight would face a much harder task of extracting
the right feature representation from the data alone.

\subsection{Training Data and Generalisation}

The \emph{training data} for both analyses consisted of the low-$z$
Cal\'an/Tololo supernovae \citep{Hamuy1996}: the $\approx 29$ events
at $z = 0.01$--$0.10$ that calibrated the nuisance parameter
$\mathcal{M}$ (and, via $H_0$, the absolute magnitude $M_B$).

The \emph{test data}---the supernovae that contribute most of the
cosmological information---were the high-$z$ events at $z = 0.3$--$0.8$.
The key ``generalisation'' step was that the MLCS or stretch
standardization learned from low-$z$ supernovae was \emph{applied
unchanged} to high-$z$ events.  Specifically:
\begin{itemize}
  \item The HZT's MLCS templates and the coefficient $\eta = 0.91$~mag
    were learned from low-$z$ training data and then used to correct
    high-$z$ events (out-of-sample prediction).
  \item The SCP's stretch factor $s$ and correction coefficient
    $\alpha_{\rm SCP}$ were similarly trained on low-$z$ data.
\end{itemize}
The implicit assumption---that Type~Ia supernovae at $z\approx 0.5$
follow the same Phillips relation as nearby events---is a
\emph{transfer learning} assumption.  Its validity was tested through
the spectroscopic comparison of high-$z$ and low-$z$ SN~Ia spectra:
both teams found that the optical spectra of high-$z$ events were,
within measurement uncertainty, identical to those of low-$z$ events
of the same light-curve shape.

The cross-$z$ universality of the shape--luminosity relation is the
supernova analogue of the cross-metal universality of the slope
$\hat\beta = h/e$ in the photoelectric problem.  Both are universality
constraints that reduce the effective degrees of freedom and make the
inference tractable with a small sample.

\subsection{Model Selection: Accelerating versus Decelerating Universes}

At the time of both papers, four main cosmological models competed:
\begin{itemize}
  \item \textbf{Einstein-de Sitter ($\Omega_M = 1,\,\Omega_\Lambda = 0$):}
    the theoretically favoured flat matter-only model.
  \item \textbf{Open matter-only ($\Omega_M \approx 0.3,\,\Omega_\Lambda =
    0$):} consistent with galaxy cluster mass measurements.
  \item \textbf{Flat $\Lambda$CDM ($\Omega_M \approx 0.3,\,\Omega_\Lambda
    \approx 0.7$):} the dark-energy model.
  \item \textbf{Flat ``QCDM'' with dynamical dark energy ($w\neq -1$):}
    generalisations where $\Omega_\Lambda$ is replaced by a field with
    equation of state $w < -1/3$.
\end{itemize}
The supernova data, combined with the flat-universe CMB evidence, delivered
decisive model selection in favour of flat $\Lambda$CDM: the best-fit
$\chi^2$ under the winning model was lower by $\Delta\chi^2 > 10$ than
under the Einstein-de Sitter alternative.  Any modern model selection
criterion---AIC, BIC, Bayesian evidence---would produce overwhelming
evidence for the dark-energy model.

\subsection{The Deductive-Inductive Loop}

Table~\ref{tab:loop} translates the supernova discovery into the language
of modern machine learning, directly paralleling the photoelectric episode
in Table~3 of \citet{Riess1998}.

\begin{table}[ht]
\centering
\caption{The hypothetico-deductive loop in the supernova episode,
translated into the language of modern machine learning.  Compare the
analogous table for the photoelectric effect in
\citep[Table~3]{Riess1998}.}
\label{tab:loop}
\begin{tabular}{p{3.6cm}p{4.0cm}p{4.0cm}}
\toprule
Step & Astrophysical name & ML name \\
\midrule
Propose Friedmann $+\Lambda$ & Theoretical model & Model architecture \\
Derive $d_L(z;\Omega_M,\Omega_\Lambda)$ & Prediction from GR &
  Hypothesis class \\
Phillips relation: use $\Delta m_{15}$, not raw $m_B$ &
  Standardization & Feature engineering \\
Cal\'an/Tololo survey & Low-$z$ calibration &
  Training data \\
High-$z$ SN Ia searches & Discovery campaigns &
  Training-data extension \\
Minimise $\chi^2(\Omega_M,\Omega_\Lambda)$ & Maximum likelihood &
  WLS training \\
$\Omega_\Lambda > 0$ at $3\sigma$ & Acceleration confirmed &
  Out-of-sample test \\
Same result from SCP and HZT & Cross-group replication &
  Zero-shot transfer \\
\bottomrule
\end{tabular}
\end{table}

%======================================================================
\section{A Deep Neural Network on the Supernova Data}
\label{sec:dnn}

\subsection{Two Hypothesis Classes Compared}

The Friedmann-equations hypothesis class~\eqref{eq:hyp_class} has three
free parameters after fixing $\sigma_{\rm int}$.  A deep neural
network (DNN) for the regression $z \to m_B$ is a universal function
approximator \citep{Hornik1989}: a three-layer network with 16 ReLU
neurons per layer has approximately
\[
  p = (1\cdot16+16) + (16\cdot16+16) + (16\cdot16+16) + (16\cdot1+1) = 593~\text{parameters}.
\]
The VC dimension of this architecture is $O(p\log p) \approx 5{,}700$
\citep{Vapnik1998}, implying that reliable generalisation requires
$n \gtrsim 5{,}700$ labeled examples.  The SCP had 42; the HZT had 50.

\subsection{Epoch I: Before Phillips (1993)}

Before the width--luminosity relation was known, the Type~Ia scatter
was $\approx 0.4$~mag.  A DNN applied to raw $(z, m_B)$ data would face
an $n=O(20)$ dataset with $\sigma_{\rm noise} \approx 0.4$~mag.

\paragraph{Underdetermination.}
With $n = 20$ and $p = 593$, the DNN is massively overparameterised.
Gradient descent finds a minimum-norm interpolant that passes through
all 20 training points but is unconstrained between them
\citep{Belkin2019}.  Applied to high-$z$ test points, it would produce
meaningless predictions.

\paragraph{Loss of the discrimination.}
The cosmological signal is $\approx 0.25$~mag at $z \approx 0.5$---
smaller than the raw $0.4$~mag scatter.  Without the Phillips feature
transformation, the DNN cannot separate signal from noise at all.

\subsection{Epoch II: After Phillips (1993), Before Large Samples}

The 18-SN Cal\'an/Tololo dataset (1996) provides clean standardized
photometry, but $n=18$ is microscopically insufficient for a 593-parameter
DNN.  The specific problems are:

\paragraph{No residual degrees of freedom.}
The Friedmann model leaves $n - 3 = 15$ residual degrees of freedom for
$n = 18$ observations and three free parameters---enough to estimate
$\hat\sigma_{\rm int}$ and compute meaningful confidence intervals.
The DNN has no such decomposition: with 593 parameters and 18 data points,
every weight and bias is undertrained; there are no residual degrees
of freedom to estimate noise variance or standard errors.

\paragraph{Cannot identify the degeneracy direction.}
The degeneracy direction $0.8\,\Omega_M - 0.6\,\Omega_\Lambda \approx
\text{const}$ (Section~\ref{sec:fisher}) must be identified from the
\emph{curvature} of $\mu(z)$ over a wide redshift range.  A DNN trained
on 18 low-$z$ events, all bunched in the near-linear regime $z < 0.1$,
has no way to estimate this curvature: the functional form is undetermined.
The Friedmann model, by contrast, predicts the curvature analytically
from the density parameters, requiring only a small number of well-placed
high-$z$ observations to pin it down.

\subsection{Epoch III: The 1998--1999 Discovery Datasets}

The SCP's 42-SN and HZT's 50-SN samples are the operative training sets
for the discovery.  What can a DNN learn from them?

\paragraph{Overfitting in the interpolation regime.}
With $n = 42$ and $p = 593$, the DNN still operates in the interpolation
regime: it can memorise all 42 training points exactly,
but its predictions between and beyond the training redshifts are
unconstrained.  In particular, the DNN has no mechanism to recover the
physically meaningful separation $\hat q_0 = -0.58$ from the overfit
weights.

\paragraph{No Cramér-Rao bound.}
The cosmological analysis produced confidence ellipses via
$\Delta\chi^2$ contours, which are valid because the Friedmann model
reduces the fit to $n - 3 = 39$ degrees of freedom.  A DNN has no
analogue: because its residuals are zero by construction (interpolation),
there is no estimate of $\sigma_{\rm int}$, no prediction interval,
and no confidence surface over $(\Omega_M, \Omega_\Lambda)$.  The DNN
\emph{cannot} reproduce the central scientific outcome, the error bars.

\paragraph{No physical interpretation of weights.}
Even if the DNN happened to learn a curve close to $\mu(z;0.28,0.72)$,
the weights of the network would not decompose into $(\hat\Omega_M,
\hat\Omega_\Lambda)$.  The connection between the learned function and
the deceleration parameter $q_0 = \Omega_M/2 - \Omega_\Lambda$ would
be entirely opaque; there is no mechanism by which the DNN could report
$\hat q_0 = -0.58$ and assign it a physical meaning.

\subsection{The Modern Era: Large Supernova Compilations}

By the 2010s, supernova compilations exceeded $n = 1{,}000$ events
(e.g., \citealt{Conley2011}, Joint Light-Curve Analysis).  In this
regime a DNN with appropriate architecture and regularisation could,
in principle, learn the shape of $\mu(z)$ from the data alone and
converge toward the Friedmann functional form.  What it still could
not do:

\begin{itemize}
  \item Identify that the learned function is $5\log_{10}\tilde d_L(z)$
    where $\tilde d_L$ satisfies a specific integral from general
    relativity.
  \item Extract $(\Omega_M, \Omega_\Lambda)$ from the network weights
    without supplying the Friedmann equations as an external constraint.
  \item Relate the curvature of $\mu(z)$ to the deceleration parameter
    $q_0 = \Omega_M/2 - \Omega_\Lambda$ and hence to the question
    of whether the universe is accelerating.
  \item Predict the $z > 1$ deceleration observed by \citet{Riess2004}:
    the DNN trained on $z < 1$ data would extrapolate the
    accelerating trend, not the decelerating one, because it has no
    physical constraint requiring the cosmological transition.
\end{itemize}

Table~\ref{tab:sample_complexity} quantifies the contrast between the
Friedmann model and a DNN at each epoch of the supernova program.

\begin{table}[ht]
\centering
\caption{Sample-complexity comparison between the three-parameter
Friedmann model (VC dimension $= 3$) and a three-layer DNN
($p = 593$ parameters, VC dim $\approx 5{,}900$).  ``Sufficient $n$''
for $\approx 1\sigma$ determination of $q_0$ at 95\% confidence is
$O(\text{VC dim}/\epsilon^2)$ for the DNN; 3 for the Friedmann model
(one more than the number of free parameters).}
\label{tab:sample_complexity}
\begin{tabular}{p{2.6cm}rp{2.4cm}p{2.4cm}p{3.8cm}}
\toprule
Epoch & $n$ available & Friedmann model & DNN
  (VC$\approx 5900$) & DNN outcome \\
\midrule
Pre-Phillips (before 1993) & $\sim 20$ raw & Insufficient$^\dagger$
  & Insufficient & Cannot detect signal (noise $> $ signal) \\
Cal\'an/ Tololo 1996 & 18 (low-$z$) & Insufficient$^\dagger$
  & Insufficient & Cannot constrain curvature of $\mu(z)$ \\
Discovery 1998 & 42--50 & Sufficient
  & Insufficient ($\times$140) & Overfits; no inference \\
JLA compilation ~2011 & $\sim 500$ & Sufficient
  & Insufficient ($\times$12) & Partially learns shape; cannot extract $q_0$ \\
Modern era 2020s & $\sim 2000$ & Sufficient
  & Borderline & Recovers $\mu(z)$ shape; cannot identify $q_0$ \\
\bottomrule
\multicolumn{5}{p{14cm}}{$^\dagger$ Without high-$z$ data, $(\Omega_M,
  \Omega_\Lambda)$ are not identified: only $\mathcal{M}$ can be
  constrained from low-$z$ observations.}
\end{tabular}
\end{table}

The lesson is precisely that of the photoelectric episode.  Einstein
reduced an infinite-dimensional function-learning problem to a
two-parameter line-fitting problem by proposing the quantum hypothesis.
Friedmann, Einstein, and the cosmological principle jointly reduced the
supernova distance-magnitude relationship to a three-parameter curve,
and Phillips reduced the scatter enough to make that curve measurable
from $\approx 40$ events.  In both cases, the theorists' contribution
was to specify the \emph{smallest useful hypothesis class}; data then
constrained the free parameters within it.

%======================================================================
\section{Conclusions}
\label{sec:conclusions}

The supernova discovery of cosmic acceleration is, at its mathematical
core, a three-parameter weighted nonlinear regression.  The free
parameters are $\mathcal{M}$ (a nuisance combination of absolute
magnitude and Hubble constant), $\Omega_M$ (matter density), and
$\Omega_\Lambda$ (cosmological constant / dark energy density).  The
measurement equation
\[
  m_{B,i} = \mathcal{M} + 5\log_{10}\tilde d_L(z_i;\Omega_M,\Omega_\Lambda)
  + \varepsilon_i
\]
is derived directly from general relativity; the Friedmann equations
are the physical hypothesis that specifies the functional form.

The central estimation challenge was not the regression itself but two
other tasks: (i)~\emph{standardizing} the supernovae using the
Phillips width--luminosity relation to reduce scatter from $\approx
0.4$~mag to $\approx 0.15$~mag (factor-of-2.7 improvement in signal-to-
noise); and (ii)~\emph{assembling} high-$z$ supernovae at
$z = 0.3$--$0.8$, where the $\mu(z)$ curve of the accelerating model
differs from that of the decelerating model by $\approx 0.2$--$0.3$~mag.

The MLE results of both groups---$\Omega_M \approx 0.28$,
$\Omega_\Lambda \approx 0.72$---give:
\begin{align*}
  \hat q_0 &= \frac{\hat\Omega_M}{2} - \hat\Omega_\Lambda
           = 0.14 - 0.72 = -0.58
           \quad (\text{acceleration at }3\sigma), \\
  z_t      &\approx 0.73
           \quad (\text{deceleration--acceleration transition}), \\
  \hat\Omega_\Lambda &= 0.72 \pm 0.09~\text{(stat)}
           \quad (\text{dark energy confirmed by both teams}).
\end{align*}

The episode illustrates three principles that echo the photoelectric story:
\begin{enumerate}
  \item \textbf{Theory specifies the hypothesis class.}  Without the
    Friedmann equations, the supernova $m_B$--$z$ relation is a
    free function of an unconstrained smooth curve.  With them, it is a
    two-parameter shape labelled by $(\Omega_M,\Omega_\Lambda)$, and
    $\approx 40$ supernovae suffice for a $3\sigma$ measurement of those
    parameters.  A purely data-driven DNN would need $O(10^3)$ examples
    to approach the same shape accuracy, and still could not extract
    $q_0$.
  \item \textbf{Feature engineering is the binding constraint.}  The
    Phillips width--luminosity relation---not the telescopes, not the
    teams' computing resources---was the key innovation that transformed
    Type~Ia supernovae from noisy candles into precision distance
    indicators.  Like Lenard's discovery that frequency (not intensity)
    controls the photoelectric stopping potential, the Phillips relation
    selected the right feature and discarded the wrong ones.
  \item \textbf{Confirming a regression curve is not the same as
    explaining dark energy.}  Riess, Schmidt, and Perlmutter confirmed
    $\Omega_\Lambda \approx 0.72$ by measuring the curvature of the
    Hubble diagram; they did not explain what dark energy is.  The
    cosmological constant remains, in Weinberg's
    \citeyearpar{Weinberg1989} phrase, one of the deepest unsolved
    problems in physics.  Millikan confirmed $h/e$ to 0.5\% without
    explaining why light consists of quanta; the supernova teams confirmed
    $\Omega_\Lambda$ to 10\% without explaining why the vacuum has the
    energy density it does.  In both cases the regression confirmed a
    number; the interpretation of that number remained---and remains---
    an open problem.
\end{enumerate}

%======================================================================
\appendix

\section{The Deceleration Parameter and the Second Accelerates Equation}
\label{app:q0}

The deceleration parameter $q_0$ follows from the second Friedmann equation.
With matter (pressure $p_M = 0$) and cosmological constant
($p_\Lambda = -\rho_\Lambda c^2$):
\begin{equation}\label{eq:friedmann2}
  \frac{\ddot a}{a} = -\frac{4\pi G}{3}\left(\rho + \frac{3p}{c^2}\right)
  + \frac{\Lambda}{3}
  = -\frac{4\pi G \rho_M}{3} + \frac{\Lambda}{3}.
\end{equation}
Evaluating at $t = t_0$ (today, $\rho_{M,0} = 3H_0^2\Omega_M/(8\pi G)$,
$\Lambda = 3H_0^2\Omega_\Lambda$):
\begin{equation}
  \frac{\ddot a_0}{a_0} = -\frac{H_0^2\Omega_M}{2} + H_0^2\Omega_\Lambda.
\end{equation}
Therefore
\begin{equation}\label{eq:q0_derive}
  q_0 = -\frac{\ddot a_0 a_0}{\dot a_0^2}
  = -\frac{\ddot a_0/a_0}{H_0^2}
  = \frac{\Omega_M}{2} - \Omega_\Lambda.
\end{equation}
For $\Omega_M < 2\Omega_\Lambda$ the universe is accelerating today.
The best-fit flat model ($\Omega_M = 0.28$, $\Omega_\Lambda = 0.72$)
gives $q_0 = 0.14 - 0.72 = -0.58$: the universe currently accelerates
at a rate $|q_0| = 0.58$ times the Hubble deceleration.

\section{Low-Redshift Expansion and the Hubble Diagram}
\label{app:lowz}

At low redshifts, the luminosity distance has a Taylor expansion in~$z$:
\begin{equation}\label{eq:dl_expand}
  d_L(z) = \frac{cz}{H_0}\left[1 + \tfrac{1}{2}(1-q_0)z + O(z^2)\right].
\end{equation}
The distance modulus then expands as
\begin{equation}\label{eq:mu_expand}
  \mu(z) = 5\log_{10}\!\left(\frac{cz}{H_0\,\text{Mpc}}\right) + 25
  + \frac{5(1-q_0)}{2\ln 10}\,z + O(z^2).
\end{equation}
At $z < 0.1$ (the Cal\'an/Tololo range), the second term is
$\approx 0.022\,(1-q_0)z < 0.004$~mag, entirely negligible compared to
$\sigma_{\rm int} \approx 0.15$~mag.  This confirms that the low-$z$
sample constrains $\mathcal{M}$ through Hubble's law but cannot
constrain $q_0$: the cosmological information is concentrated in the
second-order term, which first becomes measurable for $z \gtrsim 0.3$.

The expansion also shows that the optimal experimental design places
supernovae far into the nonlinear regime $z > 0.3$---the supernova
analogue of Millikan's insight that a wide frequency range
minimises $\operatorname{SE}(\hat\beta)$ in the photoelectric regression.

\section{The Low-Temperature Equation of State and Dark Energy}
\label{app:eos}

The cosmological constant $\Lambda$ corresponds to a fluid with equation
of state $w_\Lambda = p_\Lambda/(\rho_\Lambda c^2) = -1$: its pressure
is equal and opposite to its energy density.  A dark-energy fluid with
$w \neq -1$ (e.g., a rolling scalar field $\phi$ with potential $V(\phi)$)
also has $\ddot a > 0$ provided $w < -1/3$.

The most general measurement equation, allowing $w$ as a free parameter, is
\begin{equation}\label{eq:mu_w}
  d_L(z;\Omega_M,\Omega_{\rm de},w) = \frac{c(1+z)}{H_0}
  \int_0^z \frac{dz'}{\sqrt{\Omega_M(1+z')^3
    + \Omega_{\rm de}(1+z')^{3(1+w)}}}.
\end{equation}
The 1998--1999 analyses assumed $w = -1$ (pure $\Lambda$); later surveys
(SNLS, Union, JLA, DES) have extended the measurement to
$w = -1.028 \pm 0.032$ \citep{Conley2011,Sullivan2011}, consistent with
the cosmological constant at the 1\% level.  The discriminating power
of current supernova samples now exceeds $100\sigma$ for $w < 0$
(dark energy required), with the remaining uncertainty in $w$ lying
in the determination of systematics rather than photon statistics.

\bigskip
\bibliographystyle{plainnat}
\bibliography{riess_schmidt_perlmutter_ML}

\end{document}
